---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Exercises of Chapter 10:**
  |  **Analysis of Longitudinal Data III: Non-Normal Responses**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "22 November 2018"
output: 
  pdf_document:
    number_sections: no
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises

## Exercise 10.1

Use the epilepsy data, modifying the related R code given in the **Examples of Chapter 10**.

## Exercise 10.2

Use both the respiratory and epilepsy data sets, modifying the related R code given in the **Examples of Chapter 10**.

\pagebreak

## Exercise 10.3

See Table 10.13 in the book and use the schizophrenia2 data (see below). Modify the R codes given in the **Examples of Chapter 10** to create suitable graphics and to analyse the data.

```{r, echo=TRUE}
library("HSAUR3")
data("schizophrenia2", package = "HSAUR3")
head(schizophrenia2, 20)
```
