---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Solutions to Exercises of Chapter 7:**
  |  **Survival Analysis**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "15 November 2018"
output: 
  pdf_document:
    number_sections: no
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Solutions

## Exercise 7.2

```{r}
brcancer <- structure(list(
  time = c(23, 47, 69, 70, 100, 101, 148, 181, 198, 208, 212, 224, 5, 8, 10, 13, 18, 
           24, 26, 26, 31, 35, 40, 41, 48, 50, 59, 61, 68, 71, 76, 105, 107, 109, 
           113, 116, 118, 143, 145, 162, 188, 212, 217, 225), 
  event = c(1, 1, 1, 0, 0, 0, 1, 1, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 
            1, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0, 1, 0, 1, 1, 0, 0, 0, 0, 0, 0), 
  metastized = structure(c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 
                           2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
                           2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
                           2L, 2L, 2L, 2L, 2L), 
               .Label = c("No", "Yes"), class = "factor")),
  .Names = c("time", "event", "metastized"),
row.names = c(NA, -44L), class = "data.frame")

str(brcancer)

head(brcancer)

attach(brcancer)
library(survival)

#plot of survival function
plot(survfit(Surv(time, event) ~ metastized), xlab = "Time", ylab = "Proportion", lty = 1:2)
legend("bottomleft", c("No","Yes"), lty=1:2) 

#logrank test
survdiff(Surv(time,event) ~ metastized)

detach(brcancer)

```

\pagebreak

## Exercise 7.3

```{r}
glioma <- structure(list(
  no. = c(1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 9L, 10L, 11L, 12L, 13L, 14L, 15L, 16L,
          17L, 18L, 19L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 9L, 10L, 11L, 12L, 13L, 
          14L, 15L, 16L, 17L, 18L),
  age = c(41L, 45L, 48L, 54L, 40L, 31L, 53L, 49L, 36L, 52L, 57L, 55L, 70L, 39L, 
          40L, 47L, 58L, 40L, 36L, 27L, 32L, 53L, 46L, 33L, 19L, 32L, 70L, 72L, 
          46L, 44L, 83L, 57L, 71L, 61L, 65L, 50L, 42L), 
  sex = structure(c(1L, 1L, 2L, 2L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 2L, 1L, 1L, 
                    1L, 2L, 1L, 2L, 2L, 2L, 1L, 2L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 
                    1L, 1L, 1L, 2L, 2L, 2L, 1L), 
        .Label = c("Female", "Male"), class = "factor"),
  histology = structure(c(2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 1L, 
                          1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 
                          1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L), 
              .Label = c("GBM", "Grade3"), class = "factor"), 
  group = structure(c(2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
                      2L, 2L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
                      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L), 
          .Label = c("Control", "RIT"), class = "factor"),
  event = c(TRUE, FALSE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, FALSE, 
            FALSE, FALSE, TRUE, TRUE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, 
            TRUE, TRUE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
            TRUE, TRUE, TRUE, TRUE, TRUE), 
  time = c(53L, 28L, 69L, 58L, 54L, 25L, 51L, 61L, 57L, 57L, 50L, 43L, 20L, 14L, 
           36L, 59L, 31L, 14L, 36L, 34L, 32L, 9L, 19L, 50L, 48L, 8L, 8L, 11L, 12L, 
           15L, 5L, 8L, 8L, 6L, 14L, 13L, 25L)), 
  
  .Names = c("no.", "age", "sex", "histology", "group", "event", "time"),
  row.names = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", 
                "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", 
                "24", "25", "26", "27", "28", "29", "30", "31", "32", "33", "34", 
                "35", "36", "37"), class = "data.frame")

str(glioma)
attach(glioma)

library(survival)
glioma_cox <- coxph(Surv(time, event) ~ age + sex + histology + group)
summary(glioma_cox)
detach(glioma)

```
Required CI given in summary above - hazard on radioimmunotherapy (RIT) is between about 3% and 27% of hazard on standard therapy.

You should plot the survival curves of RIT and standard treatments.

\pagebreak

## Exercise 7.4

```{r}
teeth <- read.csv(file = "data/teeth.csv", row.names = 1)
head(teeth)
#tail(teeth)

# Set maximum longevity to 13 years and censor all at the end of the last year:
teeth$Survyear[teeth$Survyear > 13] <- 13
teeth$Reint[teeth$Survyear == 13] <- 0
```

As an example, Figure 2 of the original article (Palotie et al., 2017), entitled, "Survival (%) of restorations placed in posterior teeth (n = 5542) of 25- to 30-year-olds illustrated as medians and Kaplan-Meier curves with 95% confidence intervals by type of tooth", was created by one of the authors, *Anna Eronen*, as follows:

```{r, fig.width=8, fig.height=8}
library(survival)
#install.packages("plotrix")
library(plotrix)

plot(survfit(Surv(Survyear, Reint) ~ strata(Type2), data = teeth), conf.int = "both",
     col = c("red", "blue"), lwd = 1.7, axes = FALSE, yscale = 100,
     main = "", ylab = "", xlab = "", cex.axis = 1.2)
title(xlab = "Length of follow-up in years", line = 1.7, cex.lab = 1.2)
title(ylab = "Survivals %", line = 2.5, cex.lab = 1.2)
#
ablineclip(v=2, y1=0, y2=1.0, col="darkgrey", lty=3, lwd=1.4)
ablineclip(v=4, y1=0, y2=1.0, col="darkgrey", lty=3, lwd=1.4)
ablineclip(v=6, y1=0, y2=1.0, col="darkgrey", lty=3, lwd=1.4)
ablineclip(v=8, y1=0, y2=1.0, col="darkgrey", lty=3, lwd=1.4)
ablineclip(v=10, y1=0, y2=1.0, col="darkgrey", lty=3, lwd=1.4)
ablineclip(v=12, y1=0, y2=1.0, col="darkgrey", lty=3, lwd=1.4)
ablineclip(h=0.2, x1=0, x2=13, col="darkgrey", lty=3, lwd=1.4)
ablineclip(h=0.4, x1=0, x2=13, col="darkgrey", lty=3, lwd=1.4)
ablineclip(h=0.6, x1=0, x2=13, col="darkgrey", lty=3, lwd=1.4)
ablineclip(h=0.8, x1=0, x2=13, col="darkgrey", lty=3, lwd=1.4)
ablineclip(h=1.0, x1=0, x2=13, col="black",lty=1,lwd=1.0)
ablineclip(v=13, y1=0,y2=1,col="black",lty=1,lwd=1.0)
axis(1, at=c(0,1,3,5,7,9,11,13), labels=c(0,1,3,5,7,9,11,13), pos=0, las=1)
axis(2, at=c(0.0,0.2,0.4,0.6,0.8,1.0), labels=c(0,20,40,60,80,100), pos=0, las=2)
legend(x=0.15, y=0.15, c("Premolar","Molar"), col=c("red","blue"), lty=c(1,1), bty="n")
#
ablineclip(h=0.5, x1=0, x2=13, col="black", lty=5)
ablineclip(v=12.0, y1=0, y2=0.5, col="red", lty=5, lwd=1.7)
ablineclip(v=8.7, y1=0, y2=0.5, col="blue", lty=5, lwd=1.7)

# example of a log-rank test:
kmtest.teeth <- survdiff(Surv(Survyear, Reint) ~ Type2, data = teeth)
kmtest.teeth
```
\pagebreak
