---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Exercises of Chapter 14:**
  |  **Multidimensional Scaling and Correspondence Analysis**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "12 December 2018"
output: 
  pdf_document:
    number_sections: no
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises

## Exercise 14.2

Use the matrix gdist below, modifying the related R code given in the **Examples of Chapter 14**.

```{r}
galaxies <- c("I", "SBc", "Sc", "Sbb", "Sb", "Sba", "Sa", "SBO", "SO", "E")

g0 <- matrix(c(
0.00,0   ,0   ,0   ,0   ,0   ,0   ,0   ,0   ,0   ,
3.29,0.00,0   ,0   ,0   ,0   ,0   ,0   ,0   ,0   ,
2.79,1.13,0.00,0   ,0   ,0   ,0   ,0   ,0   ,0   ,
3.52,1.75,1.45,0.00,0   ,0   ,0   ,0   ,0   ,0   ,
3.77,2.97,1.71,2.02,0.00,0   ,0   ,0   ,0   ,0   ,
3.27,3.01,2.13,1.89,1.27,0.00,0   ,0   ,0   ,0   ,
3.93,3.72,3.00,2.25,1.86,0.68,0.00,0   ,0   ,0   ,
3.86,5.12,4.11,3.24,3.15,1.59,1.51,0.00,0   ,0   ,
3.77,5.70,4.85,3.85,3.41,1.74,2.05,0.91,0.00,0   ,
4.12,6.88,6.02,7.03,5.38,4.09,4.03,2.24,1.87,0.00)
, nrow = 10, ncol = 10, byrow = TRUE, dimnames = list(galaxies, galaxies))

gdist <- g0 + t(g0)
gdist

```
\pagebreak

## Exercise 14.3

Use the matrix P12 below, modifying the related R code given in the **Examples of Chapter 14**.

```{r}
countries <- c("Brazil", "Zaire", "Cuba", "Egypt", "France", "India", 
               "Israel", "Japan", "China", "USSR", "USA", "Yugoslavia")

    cntrs <- c("Brz", "Zai", "Cub", "Egy", "Fra", "Ind", 
               "Isr", "Jpn", "Chi", "USSR", "USA", "Yug")

P0 <- matrix(c(
0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,
4.83,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,
5.28,4.26,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,
3.44,5.00,5.17,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,
4.72,4.00,4.11,4.78,0.00,0.00,0.00,0.00,0.00,0.00,0.00,0.00,
4.50,4.83,4.00,5.83,3.44,0.00,0.00,0.00,0.00,0.00,0.00,0.00,
3.83,3.33,3.61,4.67,4.00,4.11,0.00,0.00,0.00,0.00,0.00,0.00,
3.50,3.39,2.94,9.84,4.11,4.50,4.83,0.00,0.00,0.00,0.00,0.00,
2.39,4.00,5.50,4.39,3.67,4.11,3.00,4.17,0.00,0.00,0.00,0.00,
3.06,3.39,5.44,4.39,5.06,4.50,4.17,4.61,5.72,0.00,0.00,0.00,
5.39,2.39,3.17,3.33,5.94,4.28,5.94,6.06,2.56,5.00,0.00,0.00,
3.17,3.50,5.11,4.28,4.72,4.00,4.44,4.28,5.06,6.67,3.56,0.00)
, nrow = 12, byrow = TRUE, dimnames = list(countries, cntrs))

P12 <- P0 + t(P0)
P12

```
\pagebreak

## Exercise 14.4

Use the table EyeHair below, modifying the related R code given in the **Examples of Chapter 14** to visualize and analyse the data.

**Source of the data**: Fisher, R. A. (1940). The precision of discriminant functions. *Annals of Eugenics*,
10, 422–429. (Table on p.426: "Tocher’s data for Caithness compiled by K. Maung of the Galton Laboratory"). https://doi.org/10.1111/j.1469-1809.1940.tb02264.x 

```{r}
EyeHair <- as.table(rbind(
              c(326,  38, 241, 110,  3),
              c(688, 116, 584, 188,  4),
              c(343,  84, 909, 412, 26),
              c( 98,  48, 403, 681, 85)
           ))

dimnames(EyeHair) <- list(EyeColor = c("Blue", "Light", "Medium", "Dark"),
                         HairColor = c("Fair", "Red", "Medium", "Dark", "Black"))
EyeHair
```