---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Exercises of Chapter 17:**
  |  **Cluster Analysis**
thor: "Kimmo Vehkalahti and Brian S. Everitt"
date: "19 December 2018"
output: 
  pdf_document:
    number_sections: no
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises

## Exercise 17.3

Use the matrix lowtemp below, modifying the related R code given in the **Examples of Chapter 17**.

```{r}
lowtemp <- structure(
  c(-8, -7, -44, -12, -27, 4, -25, -8, 53, 12, -22, 23, 30, -17, -6, -23, 17, -26, -16, 24, 
    0, -5, 26, 20, -12, 16, 7, 30, -2, 23, 57, 31, 6, 39, 46, 23, 12, 5, 32, 8, 13, 31, 29, 
    24, 53, 51, 35, 54, 40, 59, 43, 57, 67, 62, 36, 49, 69, 51, 52, 44, 61, 40, 33, 43, 43, 
    55, 28, 25, 5, 28, 17, 29, 3, 25, 64, 33, 11, 41, 51, 26, 28, 13, 34, 15, 8, 34, 28, 29),
.Dim = c(22L, 4L), .Dimnames = list(c("Atlanta", "Baltimore", "Bismark", "Boston", "Chicago", 
                   "Dallas", "Denver", "El Paso", "Honolulu", "Houston", "Juneau", "Los Angeles", 
                   "Miami", "Nashville", "New York", "Omaha", "Phoenix", "Portland", "Reno", 
                   "San Francisco", "Seattle", "Washington"), 
                 c("January", "April", "July", "October")))
lowtemp

```
\pagebreak

## Exercise 17.5

Use the protein data below, modifying the related R code given in the **Examples of Chapter 17**.

```{r}
protein <- read.table("data/protein.txt", sep = '\t', header = TRUE)
protein
  
```
\pagebreak

## Exercise 17.6

Use the W16F50 data below (the corresponding data of life expectancies at different ages for women), modifying the related R code given in the **Examples of Chapter 17**.

```{r}
library(dplyr)
W16F50 <- read.csv("data/w16f50.csv")
countries <- W16F50$country
row.names(W16F50) <- countries
W16F50 <- W16F50 %>% select(-country)

W16F50[c("Japan", "Italy", "Spain", "United Kingdom", "Finland", "Cuba", "United States"), ]

var(W16F50)

```
