---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Exercises of Chapter 8:**
  |  **Analysis of Longitudinal Data I: Graphical Displays and Summary Measure Approach**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "1 October 2018"
output: 
  pdf_document:
    number_sections: no
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises

## Exercise 8.1

See Table 8.2 in the book and use the BPRS data, modifying the related R code given in the **Examples of Chapter 8**.

## Exercise 8.2

Use the PAIN data, modifying the related R code given in the **Examples of Chapter 8**.

## Exercise 8.3

Use the skin resistance data (see below) and modify the R codes given in the **Examples of Chapter 8** to create suitable graphics and to analyse the data.

```{r, echo=TRUE}
library(tidyr); library(dplyr); library(ggplot2)

SKIN <- read.table("data/skin.txt", header = TRUE, sep = "\t")

glimpse(SKIN)
SKIN

```
