---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Exercises of Chapter 9:**
  |  **Analysis of Longitudinal Data II: Linear Mixed Effects Models for Normal Response Variables**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "19 November 2018"
output: 
  pdf_document:
    number_sections: no
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises

## Exercise 9.1

See Table 9.7 in the book and use the BtB data, modifying the related R code given in the **Examples of Chapter 9**.

## Exercise 9.2

Continue with the BtB data, modifying the related R code given in the **Examples of Chapter 9**.

\pagebreak

## Exercise 9.3

Use the oestrogen data (see below) and modify the R codes given in the **Examples of Chapter 9** to create suitable graphics and to analyse the data. *Obs: Instead of -9, the missing observations are already indicated by NA in the data.*

```{r, echo=TRUE}
oestrogen <- read.table("data/oestrogen.txt", sep = "\t", header = TRUE)

str(oestrogen)

# the data are readily in the long form:
head(oestrogen, n = 10); tail(oestrogen, n = 10)
```
\pagebreak

## Exercise 9.4

Use the phosphate data (see below) and modify the R codes given in the **Examples of Chapter 9** to create suitable graphics of the data and the analyze the data with an appropriate model.

```{r, echo=TRUE}
phosphate <- structure(list(

    group = structure(c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
                      1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
                      2L, 2L, 2L, 2L, 2L), 
                    .Label = c("control", "obese"), class = "factor"), 
  
  t0 = c(4.3, 3.7, 4, 3.6, 4.1, 3.8, 3.8, 4.4, 5, 3.7, 3.7, 4.4, 4.7, 4.3, 5, 
         4.6, 4.3, 3.1, 4.8, 3.7, 5.4, 3, 4.9, 4.8, 4.4, 4.9, 5.1, 4.8, 4.2, 
         6.6, 3.6, 4.5, 4.6), 
  
  t0.5 = c(3.3, 2.6, 4.1, 3, 3.8, 2.2, 3, 3.9, 4, 3.1, 2.6, 3.7, 3.1, 3.3, 4.9, 
           4.4, 3.9, 3.1, 5, 3.1, 4.7, 2.5, 5, 4.3, 4.2, 4.3, 4.1, 4.6, 3.5, 6.1, 
           3.4, 4, 4.4), 
  
  t1 = c(3, 2.6, 3.1, 2.2, 2.1, 2, 2.4, 2.8, 3.4, 2.9, 2.6, 3.1, 3.2, 3, 4.1, 
         3.9, 3.1, 3.3, 2.9, 3.3, 3.9, 2.3, 4.1, 4.7, 4.2, 4, 4.6, 4.6, 3.8,
         5.2, 3.1, 3.7, 3.8),
  
  t1.5 = c(2.6, 1.9, 2.3, 2.8, 3, 2.6, 2.5, 2.1, 3.4, 2.2, 2.3, 3.2, 3.3, 2.6, 
           3.7, 3.9, 3.1, 2.6, 2.8, 2.8, 4.1, 2.2, 3.7, 4.6, 3.4, 4, 4.1, 4.4, 
           3.6, 4.1, 2.8, 3.3, 3.8), 
  
  t2 = c(2.2, 2.9, 2.9, 2.9, 3.6, 3.8, 3.1, 3.6, 3.3, 1.5, 2.9, 3.7, 3.2, 2.2, 
         3.7, 3.7, 3.1, 2.6, 2.2, 2.9, 2.8, 2.1, 3.7, 4.7, 3.5, 3.3, 3.4, 4.1, 
         3.3, 4.3, 2.1, 2.4, 3.8), 
  
  t3 = c(2.5, 3.2, 3.1, 3.9, 3.4, 3.6, 3.4, 3.8, 3.6, 2.3, 2.2, 4.3, 4.2, 2.5,
         4.1, 4.2, 3.1, 1.9, 3.1, 3.6, 3.7, 2.6, 4.1, 3.7, 3.4, 4.1, 4.2, 4, 
         3.1, 3.8, 2.4, 2.3, 3.6), 
  
  t4 = c(3.4, 3.1, 3.9, 3.8, 3.6, 3, 3.5, 4, 4, 2.7, 3.1, 3.9, 3.7, 2.4, 4.7, 
         4.8, 3.6, 2.3, 3.5, 4.3, 3.5, 3.2, 4.7, 3.6, 3.8, 4.2, 4.4, 3.8, 3.5, 
         4.2, 2.5, 3.1, 3.8), 
  
  t5 = c(4.4, 3.9, 4, 4, 3.7, 3.5, 3.7, 3.9, 4.3, 2.8, 3.9, 4.8, 4.3, 3.4, 4.9, 
         5, 4, 2.7, 3.6, 4.4, 3.7, 3.5, 4.9, 3.9, 4, 4.3, 4.9, 3.8, 3.9, 4.8, 
         3.5, 3.3, 3.8)), 

.Names = c("group", "t0", "t0.5", "t1", "t1.5", "t2", "t3", "t4", "t5"),

row.names = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12",
              "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", 
              "24", "25", "26", "27", "28", "29", "30", "31", "32", "33"), 
class = "data.frame")
```
\pagebreak
```{r, echo=TRUE}
str(phosphate)

head(phosphate)

tail(phosphate, n = 14)
```
