---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Solutions to Exercises of Chapter 4:**
  |  **Multiple Linear Regression**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "9 October 2018"
output: 
  pdf_document:
    number_sections: no
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Solutions

## Exercise 4.1

```{r echo=TRUE}
usapol <- structure(list(

SO2 = c(10, 13, 12, 17, 56, 36, 29, 14, 10, 24, 110, 28, 17, 8, 30, 9, 47,
        35, 29, 14, 56, 14, 11, 46, 11, 23, 65, 26, 69, 61, 94, 10, 18, 9,
        10, 28, 31, 26, 29, 31, 16),

Temp = c(70.3, 61, 56.7, 51.9, 49.1, 54, 57.3, 68.4, 75.5, 61.5, 50.6, 52.3,
         49, 56.6, 55.6, 68.3, 55, 49.9, 43.5, 54.5, 55.9, 51.5, 56.8, 47.6,
         47.1, 54, 49.7, 51.5, 54.6, 50.4, 50, 61.6, 59.4, 66.2, 68.9, 51,
         59.3, 57.8, 51.1, 55.2, 45.7),

Manuf = c(213, 91, 453, 454, 412, 80, 434, 136, 207, 368, 3344, 361, 104, 125,
          291, 204, 625, 1064, 699, 381, 775, 181, 46, 44, 391, 462, 1007, 266,
          1692, 347, 343, 337, 275, 641, 721, 137, 96, 197, 379, 35, 569),

Pop = c(582, 132, 716, 515, 158, 80, 757, 529, 335, 497, 3369, 746, 201, 277,
        593, 361, 905, 1513, 744, 507, 622, 347, 244, 116, 463, 453, 751, 540,
        1950, 520, 179, 624, 448, 844, 1233, 176, 308, 299, 531, 71, 717),

Wind = c(6, 8.2, 8.7, 9, 9, 9, 9.3, 8.8, 9, 9.1, 10.4, 9.7, 11.2, 12.7, 8.3,
         8.4, 9.6, 10.1, 10.6, 10, 9.5, 10.9, 8.9, 8.8, 12.4, 7.1, 10.9, 8.6,
         9.6, 9.4, 10.6, 9.2, 7.9, 10.9, 10.8, 8.7, 10.6, 7.6, 9.4, 6.5, 11.8),

Precip = c(7.05, 48.52, 20.66, 12.95, 43.37, 40.25, 38.89, 54.47, 59.8, 48.34,
           34.44, 38.74, 30.85, 30.58, 43.11, 56.77, 41.31, 30.96, 25.94, 37,
           35.89, 30.18, 7.77, 33.36, 36.11, 39.04, 34.99, 37.01, 39.93, 36.22,
           42.75, 49.1, 46, 35.94, 48.19, 15.17, 44.68, 42.59, 38.79, 40.75, 29.07),

Days = c(36, 100, 67, 86, 127, 114, 111, 116, 128, 115, 122, 121, 103, 82, 123,
         113, 111, 129, 137, 99, 105, 98, 58, 135, 166, 132, 155, 134, 115, 147,
         125, 105, 119, 78, 103, 89, 116, 115, 164, 148, 123)),

.Names = c("SO2", "Temp", "Manuf", "Pop", "Wind", "Precip", "Days"),

row.names = c("Phoenix", "Little Rock", "San Francisco", "Denver", "Hartford",
              "Wilmington", "Washington", "Jacksonville", "Miami", "Atlanta",
              "Chicago", "Indianapolis", "Des Moines", "Wichita", "Louisville",
              "New Orleans", "Baltimore", "Detroit", "Minneapolis-St. Paul",
              "Kansas City", "St. Louis", "Omaha", "Alburquerque", "Albany",
              "Buffalo", "Cincinnati", "Cleveland", "Columbus", "Philadelphia",
              "Pittsburgh", "Providence", "Memphis", "Nashville", "Dallas",
              "Houston", "Salt Lake City", "Norfolk", "Richmond", "Seattle",
              "Charleston", "Milwaukee"), class = "data.frame")

head(usapol, n = 10)

tail(usapol, n = 10)

attach(usapol)
# Scatterplot matrix with fitted linear and locally weighted regressions
pairs(usapol, panel = function(x, y) {
                        points(x, y)
                        abline(lm(y ~ x))
                        lines(lowess(y ~ x), lty = 2)
                      }
      )

usapol_reg <- lm(SO2 ~ Temp + Manuf + Pop + Wind + Precip + Days)
summary(usapol_reg)

# Run regressions to get Rsq values
s1 <- summary(lm(Temp ~ Manuf + Pop + Wind + Precip + Days))
s2 <- summary(lm(Manuf ~ Temp + Pop + Wind + Precip + Days))
s3 <- summary(lm(Pop ~ Manuf + Temp + Wind + Precip + Days))
s4 <- summary(lm(Wind ~ Manuf + Pop + Temp + Precip + Days))
s5 <- summary(lm(Precip ~ Manuf + Pop + Wind + Temp + Days))
s6 <- summary(lm(Days ~ Manuf + Pop + Wind + Precip + Temp))
#
# VIFs
rsq <- c(s1$r.squared, s2$r.squared, s3$r.squared, s4$r.squared, s5$r.squared, s6$r.squared)
rsq
1/(1 - rsq)

# Drop population size
usapol_reg1 <- lm(SO2 ~ Temp + Manuf + Wind + Precip + Days)
step(usapol_reg1, method = "backwards")
#
# Drop Days
usapol_reg2 <- lm(SO2 ~ Temp + Manuf + Wind + Precip)
pred <- predict(usapol_reg2)
resid <- residuals(usapol_reg2)

# Normal probability plot and plot of residuals against fitted values
par(mfrow = c(1,2))
qqnorm(resid)
plot(resid ~ pred, xlab = "Fitted values", ylab = "Residuals")

# Plot of residuals against each explanatory variable in the final model
par(mfrow = c(2,2))
plot(resid ~ Temp, xlab = "Temperature", ylab = "Residuals")
plot(resid ~ Manuf, xlab="N of manuf", ylab = "Residuals")
plot(resid ~ Wind, ylab = "Residuals")
plot(resid ~ Precip, xlab = "Precipitation", ylab = "Residuals")

detach(usapol)
```
\pagebreak

## Exercise 4.2

```{r, echo=TRUE}
quality <- structure(list(

age = structure(c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 1L, 1L, 1L, 2L, 2L, 2L, 2L,
                  1L, 2L, 2L, 2L, 1L), .Label = c("A5-6", "A8-9"), class = "factor"),

sex = structure(c(1L,2L, 1L, 2L, 1L, 2L, 2L, 2L, 1L, 2L, 1L, 1L, 2L, 1L, 2L, 1L, 2L, 2L,
                  2L, 1L, 1L, 1L, 1L), .Label = c("Male", "Female"), class = "factor"),

location = structure(c(3L, 2L, 1L, 2L, 3L, 3L, 4L, 2L, 3L, 2L, 3L, 1L, 3L, 2L, 2L, 4L,
                       2L, 3L, 4L, 2L, 4L, 4L, 4L),
                     .Label = c("Home", "School", "Room", "Kroom"), class = "factor"),

coherence = c(3.81, 1.63, 3.54, 4.21, 3.3, 2.32, 4.51, 3.18, 3.02, 2.77, 3.35, 2.66,
              4.7, 4.31, 2.16, 1.89, 1.94, 2.86, 3.11, 2.9, 2.41, 2.32, 2.78),

maturity = c(3.62, 1.61, 3.63, 4.11, 3.12, 2.13, 4.31, 3.08, 3, 2.71, 3.07, 2.72,
             4.98, 4.21, 2.91, 1.87, 1.99, 2.93, 3.01, 2.87, 2.38, 2.33, 2.79),

delay = c(45, 27, 102, 39, 41, 70, 72, 41, 71, 56, 88, 13, 29, 39, 10, 15, 46, 57,
          26, 14, 45, 19, 9),

prosecute = structure(c(1L, 2L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 2L, 2L, 1L, 1L, 2L, 1L, 2L,
                        2L, 1L, 2L, 1L, 1L, 2L, 2L),
                      .Label = c("No", "Yes"), class = "factor"),

qualityct = c(34.11, 36.59, 37.23, 39.65, 42.07, 44.91, 45.23, 47.53, 54.64, 57.87,
              57.07, 45.81, 49.38, 49.53, 67.08, 83.15, 80.67, 78.47, 77.59, 76.28,
              59.64, 68.44, 65.07)),
              .Names = c("age", "sex", "location", "coherence", "maturity", "delay",
                         "prosecute", "qualityct"),

row.names = c(NA, -23L), class = "data.frame")

str(quality)
```
\pagebreak
```{r, echo=TRUE}
head(quality, n = 10); tail(quality, n = 4)

attach(quality)
plot(coherence, maturity)
cor(coherence, maturity)

quality_reg <- lm(qualityct ~ age + location + maturity + delay + prosecute)

# Show structure of dummy variables
contrasts(age)
contrasts(sex)
contrasts(location)
contrasts(prosecute)

summary(quality_reg)

step(quality_reg, method = "backwards")

quality1_reg <- lm(qualityct ~ age + delay + prosecute)

pred <- predict(quality1_reg)
resd <- residuals(quality1_reg)
```
\pagebreak
```{r, echo=TRUE, out.width = '75%'}
qqnorm(resd)
plot(pred, resd)
# etc,etc
detach(quality)
```
Maturity and coherence are highly correlated, so coherence is dropped from regression. Backwards search also drops maturity; residual plots look okay. You need to interpret the estimated regression coefficients.

\pagebreak

## Exercise 4.3

For more information, see the Wikipedia article:

https://en.wikipedia.org/wiki/Anscombe%27s_quartet

```{r, echo=TRUE}
quartet <- structure(list(
  x1 = c(10, 8, 13, 9, 11, 14, 6, 4, 12, 7, 5),
  y1 = c(8.04, 6.95, 7.58, 8.81, 8.33, 9.96, 7.24, 4.26, 10.84, 4.82, 5.68),

  x2 = c(10, 8, 13, 9, 11, 14, 6, 4, 12, 7, 5),
  y2 = c(9.14, 8.14, 8.74, 8.77, 9.26, 8.1, 6.13, 3.1, 9.13, 7.26, 4.74),

  x3 = c(10, 8, 13, 9, 11, 14, 6, 4, 12, 7, 5),
  y3 = c(7.46, 6.77, 12.74, 7.11, 7.81, 8.84, 6.08, 5.39, 8.15, 6.42, 5.73),
  
  x4 = c(8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 19),
  y4 = c(6.58, 5.76, 7.71, 8.84, 8.47, 7.04, 5.25, 5.56, 7.91, 6.89, 12.5)),

.Names = c("x1", "y1", "x2", "y2", "x3", "y3", "x4", "y4"),
row.names = c(NA, -11L), class = "data.frame"
)

quartet

attach(quartet)

# (Exceptionally) fit regressions first:

lm1 <- lm(y1 ~ x1); summary(lm1)
lm2 <- lm(y2 ~ x2); summary(lm2)
lm3 <- lm(y3 ~ x3); summary(lm3)
lm4 <- lm(y4 ~ x4); summary(lm4)
```
We get the same results on all data sets! How about some graphs?
\pagebreak
```{r, echo=TRUE}
par(mfrow = c(2,2))
plot(y1 ~ x1, xlim = c(3, 19), ylim = c(3, 13))
plot(y2 ~ x2, xlim = c(3, 19), ylim = c(3, 13))
plot(y3 ~ x3, xlim = c(3, 19), ylim = c(3, 13))
plot(y4 ~ x4, xlim = c(3, 19), ylim = c(3, 13))

pred1 <- predict(lm1)
pred2 <- predict(lm2)
pred3 <- predict(lm3)
pred4 <- predict(lm4)

resd1 <- residuals(lm1)
resd2 <- residuals(lm2)
resd3 <- residuals(lm3)
resd4 <- residuals(lm4)

plot(pred1, resd1)
plot(pred2, resd2)
plot(pred3, resd3)
plot(pred4, resd4)

detach(quartet)
```
\pagebreak

## Exercise 4.4

```{r, echo=TRUE}
fat <- structure(list(
    
  Sex = structure(c(1L, 2L, 1L, 1L, 2L, 2L, 1L, 2L, 2L, 2L,
                      2L, 1L, 2L, 2L, 2L, 1L, 2L, 2L, 2L, 2L),
                    .Label = c("F", "M"), class = "factor"),
    
    Age = c(23L, 23L, 27L, 27L, 39L, 41L, 45L, 49L, 50L, 53L,
            53L, 54L, 54L, 56L, 57L, 57L, 58L, 58L, 60L, 61L),
    
    Pcfat = c(9.5, 27.9, 7.8, 17.8, 31.4, 25.9, 27.4, 25.2, 31.1,
              34.7, 42, 20, 29.1, 32.5, 30.3, 21, 33, 33.8, 41.1, 34.5)),
    
.Names = c("Sex", "Age", "Pcfat"), row.names = c(NA, -20L ), class = "data.frame")

fat

attach(fat)
```
\pagebreak
```{r, echo=TRUE}
fat_reg <- lm(Pcfat ~ Age)
plot(Age, Pcfat, xlab = "Age", ylab = "%fat", type = "n")
text(Age, Pcfat, labels = Sex)
abline(fat_reg)

summary(lm(Pcfat ~ Age + Sex))

plot(Age, Pcfat, xlab = "Age", ylab = "%fat", type = "n")
text(Age, Pcfat, labels = Sex)
# Use figures from summary to find slope and intercepts of lines for men and women
# 5.93+11.56=17.49
abline(a = 5.93, b = 0.29)
abline(a = 17.49, b = 0.29, lty = 2)
legend("topleft", c("Females", "Males"), lty = 1:2)

summary(lm(Pcfat ~ Age * Sex))
plot(Age, Pcfat, xlab = "Age", ylab = "%fat", type = "n")
text(Age, Pcfat, labels = Sex)
abline(a = 3.47, b = 0.35)
abline(a = 20.01, b = 0.24, lty = 2)
legend("topleft", c("Females", "Males"), lty = 1:2)
detach(fat)
```
\pagebreak

## Exercise 4.5

```{r, echo=TRUE}
blood <- structure(list(
  History = structure(c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
                        1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
                        2L, 2L, 2L, 2L, 2L, 2L),
                      .Label = c("Present", "Absent"), class = "factor"),
  
  Smoking = structure(c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L,
                        3L, 3L, 3L, 3L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L,
                        3L, 3L, 3L, 3L, 3L, 3L),
                      .Label = c("NonS", "ExS", "CS"), class = "factor"),
  
  Bloodp = c(125, 156, 103, 129, 110, 128, 135, 114, 107, 134, 140, 120, 115, 135, 120,
             123, 113, 165, 145, 120, 114, 110, 91, 136, 105, 125, 103, 110, 110, 128,
             105, 90, 140, 125, 123, 108, 113, 160)),
  
.Names = c("History", "Smoking", "Bloodp"), row.names = c(NA, -38L), class = "data.frame")

str(blood)
head(blood)
tail(blood)

attach(blood)
blood_reg <- lm(Bloodp ~ History * Smoking)
summary(blood_reg)
qqnorm(residuals(blood_reg))
detach(blood)
```
\pagebreak

## Exercise 4.6

```{r, echo=TRUE}
oestrogen <- structure(list(
  Treatment = structure(c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
                          2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L),
            .Label = c("Placebo", "Active"), class = "factor"),
  
  BL1 = c(18, 25, 24, 19, 22, 27, 21, 26, 20, 24, 24, 27, 19, 25,
          19, 21, 21, 25, 25, 15, 27),
  
  BL2 = c(18, 27, 17, 15, 20, 28, 16, 26, 19, 20, 22, 27, 15, 28,
          18, 20, 21, 24, 25, 22, 26),
  
  Depression = c(15, 10, 12, 5, 5, 9, 11, 13, 6, 18, 10, 7, 8, 2,
                 6, 11, 5, 11, 6, 6, 10)),
  
.Names = c("Treatment", "BL1", "BL2", "Depression"),
row.names = c(NA, -21L), class = "data.frame")

str(oestrogen)
oestrogen

attach(oestrogen)
oestrogen_reg <- lm(Depression ~ Treatment + BL1 + BL2)
summary(oestrogen_reg)
detach(oestrogen)
#
treatCI <- c(-2.477 - 2*1.707, -2.477 + 2*1.707)
treatCI
```
CI contains the value 0, so there is no evidence of a treatment effect.
