---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Solutions to Exercises of Chapter 2:**
  |  **Looking at Data**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "8 October 2018"
output: 
  pdf_document:
    number_sections: no
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Solutions

## Exercise 2.1

## Exercise 2.2

Use the lengths data (see below).

```{r, echo=TRUE}
lengths <- structure(list(group = structure(

c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
2L, 2L, 2L, 2L, 2L, 2L),

.Label = c("Metres", "Feet"), class = "factor"),

guesses = c(8, 9, 10, 10, 10, 10, 10, 10, 11, 11, 11, 11, 12, 12, 13,
13, 13, 14, 14, 14, 15, 15, 15, 15, 15, 15, 15, 15, 16, 16, 16, 17, 17,
17, 17, 18, 18, 20, 22, 25, 27, 35, 38, 40, 24, 25, 27, 30, 30, 30, 30,
30, 30, 32, 32, 33, 34, 34, 34, 35, 35, 36, 36, 36, 37, 37, 40, 40, 40,
40, 40, 40, 40, 40, 40, 41, 41, 42, 42, 42, 42, 43, 43, 44, 44, 44, 45,
45, 45, 45, 45, 45, 46, 46, 47, 48, 48, 50, 50, 50, 51, 54, 54, 54, 55,
55, 60, 60, 63, 70, 75, 80, 94)),

.Names = c("group", "guesses"), row.names = c(NA, -113L), class = "data.frame")

head(lengths)
```
\pagebreak

```{r, echo=TRUE}
# produce side-by-side boxplots and normal probability plots
# after converting guess in meters to feet
#
attach(lengths)
feet <- group == "Feet"
convert <- ifelse(feet, 1, 3.28)
# layout(matrix(c(1,2,1,3), nrow = 2, ncol = 2, byrow = FALSE))
boxplot(I(guesses * convert) ~ group, ylab = "Width guess (in feet)", var.width = TRUE,
        names = c("Guesses in metres (after conversion to feet)", "Guesses in feet"))
```
\pagebreak

```{r, echo=TRUE}
qqnorm(guesses[!feet], ylab = "Width guesses in metres")
qqline(guesses[!feet])
```
\pagebreak

```{r, echo=TRUE}
qqnorm(guesses[feet], ylab = "Width guesses in feet")
qqline(guesses[feet])
detach(lengths)
```
\pagebreak

## Exercise 2.3

Use the states data (see below).

```{r, echo=TRUE}
states <- structure(list(

Population = c(3615, 21198, 2861, 2341, 812, 10735, 2284, 11860, 681, 472),

Income = c(3624, 5114, 4628, 3098, 4281, 4561, 4660, 4449, 4167, 3907),

Illiteracy = c(2.1, 1.1, 0.5, 2.4, 0.7, 0.8, 0.6, 1, 0.5, 0.6),

Life.Expectancy = c(69.05, 71.71, 72.56, 68.09, 71.23, 70.82, 72.13, 70.43, 72.08, 71.64),

Homicide = c(15.1, 10.3, 2.3, 12.5, 3.3, 7.4, 4.2, 6.1, 1.7, 5.5),

Graduates = c(41.3, 62.6, 59, 41, 57.6, 53.2, 60, 50.2, 52.3, 57.1),

Freezing = c(20, 20, 140, 50, 174, 124, 44, 126, 172, 168)),

.Names = c("Population", "Income", "Illiteracy", "Life.Expectancy", "Homicide",
           "Graduates", "Freezing"),

row.names = c("Alabama", "California", "Iowa", "Mississippi", "New Hampshire",
              "Ohio", "Oregon", "Pennsylvania", "South Dakota", "Vermont"),
            class = "data.frame")
states
```
\pagebreak

```{r, echo=TRUE}
# pairs plot with abbreviated state names
pairs(states, panel = function(x,y) text(x, y,
                          abbreviate(row.names(states)), cex = 0.6))
```
\pagebreak

```{r, echo=TRUE}
#install.packages("MVA")
library("MVA")

pairs(states, panel = function(x,y) {
   text(x, y, abbreviate(row.names(states)), cex = 0.6)
   par(new = T)
   bvbox(cbind(x, y), mtitle = "")
   })
```
\pagebreak

```{r, echo=TRUE}
attach(states)
coplot(Life.Expectancy ~ Homicide | Income)
detach(states)
```
\pagebreak

## Exercise 2.4

The graph commits the cardinal sin of quoting data out of context; remember that graphics often lie by omission, leaving out data sufficient for comparisons. Here, a few more data points for other years in the area would be helpful, as would similar data for other areas in which stricter enforcement of speeding had not been implemented.

## Exercise 2.5

Use the suicides data (see below).

```{r, echo=TRUE}
suicides <- structure(list(
A25.34 = c(22, 9, 22, 29, 16, 28, 48, 7, 8, 26, 4, 28, 22, 10, 20),
A35.44 = c(27, 19, 19, 40, 25, 35, 65, 8, 11, 29, 7, 41, 34, 13, 22),
A45.54 = c(31, 10, 21, 52, 36, 41, 84, 11, 18, 36, 10, 46, 41, 15, 28),
A55.64 = c(34, 14, 31, 53, 47, 49, 81, 18, 20, 32, 16, 51, 50, 17, 33),
A65.74 = c(24, 27, 49, 69, 56, 52, 107, 27, 28, 28, 22, 35, 51, 22, 37)),

.Names = c("A25.34", "A35.44", "A45.54", "A55.64", "A65.74"),

row.names = c("Canada", "Israel", "Japan", "Austria", "France", "Germany",
              "Hungary","Italy", "Netherlands", "Poland", "Spain", "Sweden",
              "Switzerland", "UK", "USA"),
            class = "data.frame")
suicides
```
\pagebreak

```{r, echo=TRUE}
boxplot(suicides[, 1], suicides[, 2], suicides[, 3], suicides[, 4], suicides[, 5],
        names = c("25-34", "35-44", "45-54", "55-64", "65-74"),
        ylab = "Male suicide rates per 100,000")
```

A set of boxplots for the different countries might also be interesting.
