---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Exercises of Chapter 7:**
  |  **Survival Analysis**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "15 November 2018"
output: 
  pdf_document:
    number_sections: no
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises

## Exercise 7.2

Use the brcancer data (see below) and modify the R code given in the **Examples of Chapter 7** to analyse and visualize the data as instructed.

```{r, echo=TRUE}
brcancer <- structure(list(
  time = c(23, 47, 69, 70, 100, 101, 148, 181, 198, 208, 212, 224, 5, 8, 10, 13, 18, 
           24, 26, 26, 31, 35, 40, 41, 48, 50, 59, 61, 68, 71, 76, 105, 107, 109, 
           113, 116, 118, 143, 145, 162, 188, 212, 217, 225), 
  event = c(1, 1, 1, 0, 0, 0, 1, 1, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 
            1, 1, 1, 1, 1, 1, 1, 0, 0, 0, 0, 1, 0, 1, 1, 0, 0, 0, 0, 0, 0), 
  metastized = structure(c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 
                           2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
                           2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
                           2L, 2L, 2L, 2L, 2L), 
               .Label = c("No", "Yes"), class = "factor")),
  .Names = c("time", "event", "metastized"),
row.names = c(NA, -44L), class = "data.frame")

str(brcancer)

head(brcancer)
```
\pagebreak

## Exercise 7.3

Use the glioma data (see below) and modify the R code given in the **Examples of Chapter 7** to analyse the data as instructed.

```{r, echo=TRUE}
glioma <- structure(list(
  no. = c(1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 9L, 10L, 11L, 12L, 13L, 14L, 15L, 16L,
          17L, 18L, 19L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 9L, 10L, 11L, 12L, 13L, 
          14L, 15L, 16L, 17L, 18L),
  age = c(41L, 45L, 48L, 54L, 40L, 31L, 53L, 49L, 36L, 52L, 57L, 55L, 70L, 39L, 
          40L, 47L, 58L, 40L, 36L, 27L, 32L, 53L, 46L, 33L, 19L, 32L, 70L, 72L, 
          46L, 44L, 83L, 57L, 71L, 61L, 65L, 50L, 42L), 
  sex = structure(c(1L, 1L, 2L, 2L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 2L, 1L, 1L, 
                    1L, 2L, 1L, 2L, 2L, 2L, 1L, 2L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 
                    1L, 1L, 1L, 2L, 2L, 2L, 1L), 
        .Label = c("Female", "Male"), class = "factor"),
  histology = structure(c(2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 1L, 
                          1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 
                          1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L), 
              .Label = c("GBM", "Grade3"), class = "factor"), 
  group = structure(c(2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
                      2L, 2L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
                      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L), 
          .Label = c("Control", "RIT"), class = "factor"),
  event = c(TRUE, FALSE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, FALSE, 
            FALSE, FALSE, TRUE, TRUE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, 
            TRUE, TRUE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
            TRUE, TRUE, TRUE, TRUE, TRUE), 
  time = c(53L, 28L, 69L, 58L, 54L, 25L, 51L, 61L, 57L, 57L, 50L, 43L, 20L, 14L, 
           36L, 59L, 31L, 14L, 36L, 34L, 32L, 9L, 19L, 50L, 48L, 8L, 8L, 11L, 12L, 
           15L, 5L, 8L, 8L, 6L, 14L, 13L, 25L)), 
  
  .Names = c("no.", "age", "sex", "histology", "group", "event", "time"),
  row.names = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", 
                "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", 
                "24", "25", "26", "27", "28", "29", "30", "31", "32", "33", "34", 
                "35", "36", "37"), class = "data.frame")

str(glioma)
```
\pagebreak

## Exercise 7.4

**Source of the data**: Palotie, U., Eronen, A. K., Vehkalahti, K. and Vehkalahti, M. M. (2017). Longevity of 2- and 3-surface restorations in posterior teeth of 25- to 30-year-olds attending Public Dental Service—A 13-year observation. *Journal of Dentistry*, 62, 13–17. https://doi.org/10.1016/j.jdent.2017.05.012

<<<<<<< HEAD
(Acknowledgements, p.17: "The authors acknowledge the Helsinki City Public Dental Service for handing their data at disposal. *Tuomo Maisala* is thanked for picking out the cases and *Marjut Grainger* for invaluable help with data cleaning.")
=======
The authors acknowledge the Helsinki City PDS for handing their data at disposal. *Tuomo Maisala* is thanked for picking out the cases and *Marjut Grainger* for invaluable help with data cleaning.
>>>>>>> 188fa7aa4223e3810edde200a468b3f9066f2b77

```{r, echo=TRUE}
teeth <- read.csv(file = "data/teeth.csv", row.names = 1)
head(teeth)
tail(teeth)

# Set maximum longevity to 13 years and censor all at the end of the last year:
teeth$Survyear[teeth$Survyear > 13] <- 13
teeth$Reint[teeth$Survyear == 13] <- 0
```

- ID: Tooth ID
- Gender: 1=Male, 2=Female
- Age: Patient age in years (2002)
- ISOcode: Tooth code (ISO)
- Reint: Re-intervention: 0=No, 1=Yes
- Survyear: Longevity of restoration in years
- Type4: 1=Upper premolar, 2=Lower premolar, 3=Upper molar, 4=Lower molar
- Material: 1=Amalgam, 2=Composite
- Type2: 1=Premolar, 2=Molar
