---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Examples of Chapter 17:**
  |  **Cluster Analysis**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "18 December 2018"
output: 
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak

# Examples

# Table 12.1: Chest, Waist, and Hip Measurements of 20 Individuals

This data set was introduced in **Chapter 12** and it is briefly revisited here.

```{r}
body <- structure(list(
  Chest = c(34, 37, 38, 36, 38, 43, 40, 38, 40, 41, 36, 36, 34, 33, 36, 37, 34, 36, 38, 35), 
  Waist = c(30, 32, 30, 33, 29, 32, 33, 30, 30, 32, 24, 25, 24, 22, 26, 26, 25, 26, 28, 23), 
  Hips = c(32, 37, 36, 39, 33, 38, 42, 40, 37, 39, 35, 37, 37, 34, 38, 37, 38, 37, 40, 35)), 
  .Names = c("Chest", "Waist", "Hips"), row.names = c(NA, -20L), class = "data.frame")
body
```
\pagebreak

## Figure 17.5

```{r, out.width='100%', fig.width=6, fig.height=6}
attach(body)
distances <- dist(body)

body_sl3 <- cutree(hclust(distances, method = "single"), h=3.8)
body_cl2 <- cutree(hclust(distances, method = "complete"), h=10)
body_al2 <- cutree(hclust(distances, method = "average"), h=7.8)

layout(matrix(c(1,2,3,4,5,6), 2, 3, byrow=TRUE), c(1,1,1), c(2,1), TRUE)

plot(hclust(distances, method = "single"), ylab =" Height", sub = "Single linkage")
plot(hclust(distances, method = "complete"), ylab =" Height", sub = "Complete linkage")
plot(hclust(distances, method = "average"), ylab = "Height", sub = "Average linkage")

body_pc <- princomp(body)
xlim <- range(body_pc$scores[, 1])

plot(body_pc$scores[, 1:2], type = "n",xlim = xlim, ylim =xlim)
text(body_pc$scores[, 1:2], labels = body_sl3, cex=0.8)

plot(body_pc$scores[, 1:2], type = "n",xlim = xlim, ylim = xlim)
text(body_pc$scores[, 1:2], labels = body_cl2, cex=0.8)

plot(body_pc$scores[, 1:2], type = "n", xlim = xlim, ylim = xlim)
text(body_pc$scores[, 1:2], labels = body_al2, cex=0.8)

detach(body)
```
\pagebreak

# Table 17.1: Life Expectancies at Different Ages for Men in Seven Countries

```{r}
library(dplyr)
W16M50 <- read.csv("data/w16m50.csv")
countries <- W16M50$country
row.names(W16M50) <- countries
W16M50 <- W16M50 %>% select(-country)

W16M50[c("Japan", "Italy", "Spain", "United Kingdom", "Finland", "Cuba", "United States"), ]

var(W16M50)
```
\pagebreak

## Figure 17.6

```{r}
distances <- dist(W16M50)
hclu <- hclust(distances, method = "complete")

#install.packages("ggdendro")
library(ggdendro)
library(ggplot2)

p1 <- ggdendrogram(hclu, rotate = TRUE, theme_dendro = FALSE)
p2 <- p1 + theme_bw()
p3 <- p2 + scale_y_continuous(name = "Height", breaks = seq(0, 25, 5)) # obs: rotation!
p4 <- p3 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p5 <- p4 + xlab("") # obs: rotation!
p6 <- p5 + theme(axis.text.y = element_text(color = "black", size = 7)) # obs: rotation!
p6
```
\pagebreak

## Figure 17.7

```{r, fig.width=6, fig.height=6}
life_cl4 <- cutree(hclu, k=4)
pairs(W16M50, panel = function(x,y) text(x, y, labels = life_cl4, cex=1.0))

```
\pagebreak

## Figure 17.8

```{r, out.width='100%'}
life_pc <- princomp(W16M50, cor = TRUE)
pc_scores <- as.data.frame(life_pc$scores)
pc_scores <- cbind(pc_scores, life_cl4)
pc_scores$life_cl4 <- as.factor(pc_scores$life_cl4)

pcvar <- round(100 * life_pc$sdev / sum(life_pc$sdev), 2)
xlabel <- paste("PC 1 (", pcvar[1], " %)", sep = "")
ylabel <- paste("PC 2 (", pcvar[2], " %)", sep = "")

p1 <- ggplot(pc_scores, aes(x = Comp.1, y = Comp.2, colour = life_cl4))
# p1 <- ggplot(pc_scores, aes(x = Comp.1, y = Comp.2)) # b&w points (book)
p2 <- p1 + geom_text(aes(label = life_cl4), size=3)
p3 <- p2 + geom_text(aes(label = countries), position = position_nudge(y = -0.15), size=1.5)
p4 <- p3 + scale_x_continuous(name = xlabel, 
                              breaks = seq(-4, 3, 1)) + scale_y_continuous(name = ylabel)
p5 <- p4 + theme_bw()
p6 <- p5 + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                 legend.position = "none")
p7 <- p6 + coord_fixed(ratio = 1)
p7

```
\pagebreak

## Table 17.2

```{r}
country.mean <- lapply(1:4, function(nc) apply(W16M50[life_cl4 == nc, ], 2, mean))
country.clus <- lapply(1:4, function(nc) countries[life_cl4 == nc])

country.mean
country.clus
```
\pagebreak

# Figure 17.9: Plot of within-groups sum of squares against number of clusters

See **Chapter 13**, where this data set was used for the first time.

```{r}
crime <- read.table("data/crime.txt", sep = '\t')
rlabs <- row.names(crime)

head(crime)
# DC (outlier, see Chapter 13):
crime[24, ]

# remove DC:
crime <- crime[-24, ]

# variances:
apply(crime, 2, var)

# standardize by range
rge <- apply(crime, 2, max) - apply(crime, 2, min)
crime_std <- sweep(crime, 2, rge, FUN = "/")

# variances of the std data:
apply(crime_std, 2, var)
```
\pagebreak
```{r}
# plot of wgss against number of clusters:
n <- length(crime_std[, 1])
wss1 <- (n-1) * sum(apply(crime_std, 2, var))
wss <- numeric(0)
for(i in 2:6) {
    W <- sum(kmeans(crime_std, i)$withinss)
    wss <- c(wss, W)
}
wss <- c(wss1, wss)
plot(1:6, wss, type = "l", 
     xlab = "Number of groups", ylab = "Within groups sum of squares", lwd=2)

```
\pagebreak

## Table 17.3

```{r}
# get two-group solution from k-means and group means and membership
crime_kmean2 <- kmeans(crime_std, 2)
lapply(1:2, function(nc) apply(crime[crime_kmean2$cluster == nc, ], 2, mean))
lapply(1:2, function(nc) rlabs[crime_kmean2$cluster == nc])
```
\pagebreak

## Figure 17.10

```{r}
crime_pc <- princomp(crime_std, cor = TRUE)
xlim <- range(crime_pc$scores[, 1])
plot(crime_pc$scores[, 1:2], type = "n", xlim = xlim, ylim = xlim)
text(crime_pc$scores[, 1:2], labels = crime_kmean2$cluster, cex=0.8)

```

\pagebreak

# Table 17.4: Proportion of Respondents Answering Yes to Each of the Questions in the Survey of Gastroenterologists

```{r}
prop <- read.table("data/prop.txt", sep = '\t')
options(digits=3)
prop
```
\pagebreak

## Figure 17.11

```{r}
#install.packages("mclust")
library(mclust)

# obs: seeking a correspondence with the 1st ed. (2009) version that was done
# using mclust ver.2 (whereas the results below were achieved with mclust ver.5):
prop_mclust <- Mclust(prop, modelNames = c("EII", "VII", "EEI", "VVI", "EEE", "VVV"))
plot(prop_mclust, what = "BIC", symbols = c("5","2","1","6","3","4"), 
     xlab = "Number of Clusters", ylim = c(-110,50))
# Note:
# "EII" = spherical, equal volume
# "VII" = spherical, unequal volume
# "EEI" = diagonal, equal volume and shape
# "VVI" = diagonal, varying volume and shape
# "EEE" = ellipsoidal, equal volume, shape, and orientation
# "VVV" = ellipsoidal, varying volume, shape, and orientation
# (there are several more options in mclust ver.5)

```
\pagebreak

## Table 17.5

```{r}
# obs: the optimal model (4: "VVV") with mclust ver.5 has only two clusters
prop_mclust$parameters$mean[, 1]
prop_mclust$parameters$mean[, 2]
row.names(prop)[prop_mclust$classification == 1]
row.names(prop)[prop_mclust$classification == 2]
```
