---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Examples of Chapter 10:**
  |  **Analysis of Longitudinal Data III: Non-Normal Responses**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "20 November 2018"
output: 
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak

# Examples

ENSIN: T 10.3 ja 10.4: using BtB data!

vasta sitten:

# Table 10.1: Respiratory Disorder Data

```{r, echo=TRUE}
```
\pagebreak

(tähän väliin Fig. 10.2 Tables 10.5, 10.6, 10.7, 10.10) - all with respiratory data

# Table 10.2: Data from a Clinical Trial of Patients Suffering from Epilepsy

(tän jälkeen Fig 10.3, Tables 10.8, 10.9, 10.11, 10.12) - all with epilepsy data

```{r, echo=TRUE}
```

wanhoja pohjia ch9:Stä:


\pagebreak

## Figure 9.1

```{r, echo=TRUE}
```
\pagebreak

## Table 9.3

```{r, echo=TRUE}
```
\pagebreak

## Figure 9.2

```{r, echo=TRUE}
```
\pagebreak

## Figure 9.3

```{r, echo=TRUE, fig.width=10, fig.height=10}
```
\pagebreak

## Table 9.4

```{r, echo=TRUE}
```
\pagebreak

## Table 9.5

```{r, echo=TRUE}
```
\pagebreak

## Table 9.6

```{r, echo=TRUE}
```
\pagebreak

## Figure 9.4

```{r, echo=TRUE, fig.width=3, fig.height=4}

```
\pagebreak

# Table 9.7: Patients in Each Treatment Group of the “Beat the Blues” (BtB) Clinical Trial of CBT for Depression

```{r, echo=TRUE}
library(tidyr); library(dplyr); library(ggplot2)

BtB <- read.table("data/BtB.txt", header = TRUE, sep = '\t')

# to make sure that the factor levels are logical (esp. Treatment):
BtB <- within(BtB, {
       Drug <- factor(Drug, levels=c("No", "Yes")) # default
     Length <- factor(Length, levels=c("<6m", ">6m")) # default
  Treatment <- factor(Treatment, levels=c("TAU", "BtheB")) # NOT default!
})

glimpse(BtB); head(BtB); tail(BtB)

```
\pagebreak

## Figure 9.5

```{r, echo=TRUE}
# Convert data to long form, including the baseline BDI measurement:
BtBL0 <- gather(BtB, key = Visit, value = BDI, BDIpre, BDI2m, BDI4m, BDI6m, BDI8m)

glimpse(BtBL0); head(BtBL0); tail(BtBL0)

p1 <- ggplot(BtBL0, aes(x = factor(Visit), y = BDI, fill = Treatment))
p2 <- p1 + geom_boxplot()
p3 <- p2 + facet_grid(Treatment ~., labeller = label_parsed)
p4 <- p3 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p5 <- p4 + theme(legend.position = "none")
p6 <- p5 + scale_x_discrete(name = "", limits = c("BDIpre", "BDI2m", "BDI4m", "BDI6m", "BDI8m")) # BDIpre first!
p6
# p7 <- p6 + scale_fill_grey(start = 1, end = 1) # B&W in the book
# p7
```
\pagebreak

## Figure 9.6

```{r, echo=TRUE, fig.width=10, fig.height=10}
pairs(BtB[, -c(1:4)], cex = 0.8, cex.labels = 1.0)
```
\pagebreak

## Table 9.8

```{r, echo=TRUE}
# Convert data to long form for the analyses, adding Time:
BtBL <- gather(BtB, key = Visit, value = BDI, BDI2m, BDI4m, BDI6m, BDI8m) %>%
  mutate(Time = as.integer(substr(Visit, 4, 4))) 

glimpse(BtBL); head(BtBL); tail(BtBL)

BtB_fit0 <- lm(BDI ~ BDIpre + Time + Treatment + Drug + Length, data = BtBL, na.action = na.omit)
summary(BtB_fit0)
```
\pagebreak

## Table 9.9

```{r, echo=TRUE}
library("lme4")

BtB_fit1 <- lmer(BDI ~ BDIpre + Time + Treatment + Drug + Length + (1 | Subject), 
                 data = BtBL, na.action = na.omit)
summary(BtB_fit1)
```
\pagebreak
```{r, echo=TRUE}
BtB_fit2 <- lmer(BDI ~ BDIpre + Time + Treatment + Drug + Length + (Time | Subject), 
                 data = BtBL, na.action = na.omit)
summary(BtB_fit2)

anova(BtB_fit2, BtB_fit1)
```
