---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Examples of Chapter 13:**
  |  **Principal Components Analysis**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "5 December 2018"
output: 
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak

# Examples

# Table 13.1: Correlations of Blood Chemistry Variables and Their Standard Deviations

```{r, comment=''}
blood_chem <- c(1.000, 0.290, 0.202,-0.055,-0.105,-0.252,-0.229, 0.058,
                0.290, 1.000, 0.415, 0.285,-0.376,-0.349,-0.164,-0.129,
                0.202, 0.415, 1.000, 0.419,-0.521,-0.441,-0.145,-0.076,
               -0.055, 0.285, 0.419, 1.000,-0.877,-0.076, 0.023,-0.131,
               -0.105,-0.376,-0.521,-0.877, 1.000, 0.206, 0.034, 0.151,
               -0.252,-0.349,-0.441,-0.076, 0.206, 1.000, 0.192, 0.077,
               -0.229,-0.164,-0.145, 0.023, 0.034, 0.192, 1.000, 0.423,
                0.058,-0.129,-0.076,-0.131, 0.151, 0.077, 0.423, 1.000)
blood_corr <- matrix(blood_chem, ncol = 8)
sds <- c(0.371, 41.253, 1.935, 0.077, 0.071, 4.037, 2.732, 0.297)
names <- c("rBlood", "Plate", "wBlood", "Neut.", "Lymph", "Bilir.", "Sodium", "Potass.")
dimnames(blood_corr) <- list(names, names)

blood_corr
sds
```
\pagebreak

## Table 13.2

```{r}
# Calculate the covariance matrix:
blood_cov <- diag(sds) %*% blood_corr %*% diag(sds)
dimnames(blood_cov) <-list(names, names)
blood_pcacov <- princomp(covmat = blood_cov)

summary(blood_pcacov, loadings = TRUE)
```
\pagebreak

## Table 13.3

```{r}
blood_pcacor <- princomp(covmat = blood_corr)
summary(blood_pcacor, loadings = TRUE)
```
\pagebreak

## Figure 13.1

```{r, fig.width=7, fig.height=7}
par(mfrow = c(2,1))
plot(1:8, blood_pcacor$sdev^2, 
     xlab = "Component number", ylab = "Component variance", type = "l")
title("Scree Diagram")

plot(1:8, log(blood_pcacor$sdev^2), 
     xlab = "Component number", ylab = "log(component variance)", type = "l")
title("Log(Eigenvalue) Diagram")
```
\pagebreak

# Table 13.4: Head Lengths (in Millimeters) of First and Second Sons of 25 Families

```{r}
head_dat <- structure(
  c(191, 195, 181, 183, 176, 208, 189, 197, 188, 192, 179, 183, 
    174, 190, 188, 163, 195, 186, 181, 175, 192, 174, 176, 197, 190, 
    179, 201, 185, 188, 171, 192, 190, 189, 197, 187, 186, 174, 
    185, 195, 187, 161, 183, 173, 182, 165, 185, 178, 176, 200, 187), 
  .Dim = c(25L, 2L), .Dimnames = list(NULL, c("HLFS", "HLSS")))
head_dat

# get principal components
head_pc <- princomp(covmat = var(head_dat))
head_pc

print(summary(head_pc), digits = 3, loadings = TRUE)

# pc scores for first family
head_pc1 <- princomp(head_dat)
head_pc1$scores[1, ]

# check
0.693*(191-185.72)+0.721*(179-183.84)
-0.721*(191-185.72)+0.693*(179-183.84)

y1 <- 12.9524588 * c(0.692986, 0.720951)
y2 <- 5.3229513 * c(-0.720951, 0.692986)
X <- cbind(y1, y2)
X %*% t(X)

```
\pagebreak

## Figure 13.2

```{r, fig.width=7, fig.height=7}
a1 <- 183.84-0.721*185.72/0.693
b1 <- 0.721/0.693
a2 <- 183.84-(-0.693*185.72/0.721)
b2 <- -0.693/0.721
plot(head_dat, xlab = "First Son's Head Length (mm)", ylab = "Second Son's Head Length")
abline(a1, b1)
abline(a2, b2, lty=2)
```
\pagebreak

## Figure 13.3

```{r, fig.width=7, fig.height=7}
xlim <- range(head_pc1$scores[, 1])
plot(head_pc1$scores, xlim = xlim, ylim = xlim)
```
\pagebreak

# Table 13.5: Crime Rates in the United States

```{r}
crime <- read.table("data/crime.txt", sep = '\t')
str(crime)

crime
```
\pagebreak

## Figure 13.4

```{r, fig.width=10, fig.height=10}
pairs(crime)
```
\pagebreak

## Table 13.6

```{r}
crime_pc <- princomp(crime, cor = TRUE)
summary(crime_pc, loadings = TRUE)
```
\pagebreak

## Figure 13.5

```{r}
plot(1:7, crime_pc$sdev^2, type = "l", xlab = "Component Number", ylab = "Variance")
```
\pagebreak

## Figure 13.6

```{r, fig.width=7, fig.height=7}
xlim <- range(crime_pc$scores[, 1])
plot(crime_pc$scores[, 1], crime_pc$scores[, 2],
     xlab = "First PC score", ylab = "Second PC score", xlim = xlim, ylim = xlim, type = "n")
text(crime_pc$scores[, 1], crime_pc$scores[, 2], row.names(crime), cex=1.0) 
```
\pagebreak

## Table 13.7

```{r}
options(digits=3)
# use all pcas to get original correlation matrix
lambda <- crime_pc$sdev^2
Astar <- crime_pc$loadings[1:7, 1:7]
R <- Astar %*% diag(lambda) %*% t(Astar)
R

# predicted correlation matrix based on first two components
lambda2 <- lambda[1:2]
Astar2 <- Astar[,1:2]
R2 <- Astar2 %*% diag(lambda2) %*% t(Astar2)
R2
```
\pagebreak

# Table 13.8: Correlation Matrix for Drug Usage Data

```{r}
drugs_corr <- structure(c(
 
 1,     0.447, 0.442, 0.435, 0.114, 0.203, 0.091, 0.082, 0.513, 0.304, 0.245, 0.101, 0.245, 
 0.447, 1,     0.619, 0.604, 0.068, 0.146, 0.103, 0.063, 0.445, 0.318, 0.203, 0.088, 0.199, 
 0.442, 0.619, 1,     0.583, 0.053, 0.139, 0.110, 0.066, 0.365, 0.240, 0.183, 0.074, 0.184, 
 0.435, 0.605, 0.583, 1,     0.115, 0.258, 0.122, 0.097, 0.482, 0.368, 0.255, 0.139, 0.293, 
 0.114, 0.068, 0.053, 0.115, 1,     0.349, 0.209, 0.321, 0.186, 0.303, 0.272, 0.279, 0.278, 
 0.203, 0.146, 0.139, 0.258, 0.349, 1,     0.221, 0.355, 0.315, 0.377, 0.323, 0.367, 0.545, 
 0.091, 0.103, 0.110, 0.122, 0.209, 0.221, 1,     0.201, 0.150, 0.163, 0.310, 0.232, 0.232, 
 0.082, 0.063, 0.066, 0.097, 0.321, 0.355, 0.201, 1,     0.154, 0.219, 0.288, 0.320, 0.314, 
 0.513, 0.445, 0.365, 0.482, 0.186, 0.315, 0.150, 0.154, 1,     0.534, 0.301, 0.204, 0.394, 
 0.304, 0.318, 0.240, 0.368, 0.303, 0.377, 0.163, 0.219, 0.534, 1,     0.302, 0.368, 0.467, 
 0.245, 0.203, 0.183, 0.255, 0.272, 0.323, 0.310, 0.288, 0.301, 0.302, 1,     0.304, 0.392, 
 0.101, 0.088, 0.074, 0.139, 0.279, 0.367, 0.232, 0.320, 0.204, 0.368, 0.304, 1,     0.511, 
 0.245, 0.199, 0.184, 0.293, 0.278, 0.545, 0.232, 0.314, 0.394, 0.467, 0.392, 0.511, 1    

 ), .Dim = c(13L, 13L))

drugs <- c("Cigarettes", "Beer", "Wine", "Liquor", "Cocaine", "Tranquilizers", "Drugstore",
           "Heroin", "Marijuana", "Hashish", "Inhalants", "Hallucinogenics", "Amphetamine")
dimnames(drugs_corr) <- list(drugs, c(1:13))
drugs_corr
```
\pagebreak

## Table 13.9

```{r}
drugs_pc <- princomp(covmat = drugs_corr)
print(drugs_pc)
summary(drugs_pc, loadings=TRUE)
```
\pagebreak

## Figure 13.7

```{r}
plot(1:13, drugs_pc$sdev^2, type = "l", xlab = "Component Number", ylab = "Variance")
```
