---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Examples of Chapter 7:**
  |  **Survival Analysis**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "14 November 2018"
output: 
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak

# Examples

# Table 7.1: Age at First Sexual Intercourse for Women in Two Diagnostic Groups

```{r, echo=TRUE}
first_sex <- structure(list(
  Diagnosis = structure(c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
                          1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
                          1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
                          2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
                          2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
                          2L, 2L, 2L, 2L), .Label = c("RAN", "NC"), class = "factor"),
  
  Agefs = c(30, 24, 12, 21, 19, 18, 24, 20, 24, 28, 17, 18, 18, 27, 31, 17, 28, 29, 23,
            19, 19, 18, 26, 22, 20, 28, 20, 21, 18, 18, 20, 21, 17, 21, 16, 16, 18, 21,
            17, 17, 15, 13, 15, 18, 16, 19, 14, 16, 17, 16, 17, 18, 16, 16, 16, 17, 16, 
            19, 19, 17, 18, 20, 16, 16, 17, 17, 17, 18, 16, 16, 19, 19, 18, 17, 19, 16,
            16, 15, 19, 20, 16, 15, 17, 18),
  
  Age = c(30, 24, 18, 42, 19, 39, 30, 30, 33, 38, 17, 18, 18, 27, 31, 17, 28, 29, 23,
          35, 28, 18, 26, 22, 20, 28, 26, 21, 22, 18, 25, 21, 17, 21, 22, 20, 21, 29, 
          20, 20, 20, 20, 20, 20, 19, 20, 20, 20, 20, 21, 20, 22, 22, 20, 38, 21, 21, 
          22, 36, 24, 30, 39, 20, 19, 22, 22, 23, 18, 29, 19, 22, 22, 21, 19, 21, 20, 
          22, 18, 26, 23, 20, 21, 21, 21), 
  
  Status = c(0, 0, 1, 1, 0, 1, 1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 0, 0, 0, 
             0, 1, 0, 1, 0, 1, 0, 0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 
             1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 
             1, 1, 1, 1, 1, 1, 1, 1, 1)), 
  
.Names = c("Diagnosis", "Agefs", "Age", "Status"),
row.names = c(NA, -84L), class = "data.frame")

head(first_sex)
tail(first_sex)
```
\pagebreak

## Table 7.2

```{r, echo=TRUE}
attach(first_sex)
library("survival")
summary(survfit(Surv(Agefs, Status) ~ 1))
detach(first_sex)
```
\pagebreak

## Figure 7.1

```{r, echo=TRUE}
attach(first_sex)
plot(survfit(Surv(Agefs, Status) ~ 1), xlab = "Age at first sex (years)", ylab = "Proportion")
legend("bottomleft", c("Estimated survival function", "Confidence interval (95%)"), lty = 1:2)
detach(first_sex)
```
\pagebreak

## Figure 7.2

The following gives also the log-rank test and Cox regression results that are referred to in the text.

```{r, echo=TRUE}
attach(first_sex)
plot(survfit(Surv(Agefs,Status) ~ Diagnosis), xlab = "Age at first sex (years)",
     ylab = "Proportion", lty = 1:2)
legend("bottomleft", c("RAN", "NC"), lty = 1:2)

# logrank test
survdiff(Surv(Agefs, Status) ~ Diagnosis)
#cox regression
sex_cox <- coxph(Surv(Agefs, Status) ~ Diagnosis)
summary(sex_cox)

contrasts(Diagnosis)
detach(first_sex)
```
\pagebreak

# Table 7.3: Data for Heroin Addicts Being Treated with Methadone

```{r, echo=TRUE}
heroin_dat <- structure(list(
  Clinic = structure(c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
                       1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
                       1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
                       1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
                       1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
                       1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
                       1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
                       2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
                       2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
                       2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
                       2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 1L, 1L,
                       2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
                       1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
                       2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 
                       2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 
                       1L, 1L, 1L, 2L, 1L, 2L, 2L, 1L, 1L, 2L, 2L, 1L, 1L), 
                     .Label = c("Clinic 1", "Clinic 2"), class = "factor"), 
  
  Status = c(1, 1, 1, 1, 1, 1, 1, 0, 1, 1,0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 
             1, 0, 1, 1, 0, 1, 1, 0, 1, 1, 1, 0, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 1, 1, 
             1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 0, 1, 0, 1, 0, 1, 0, 1, 1, 
             1, 0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 0, 0, 1, 1, 0, 1, 0, 1, 1, 1, 1, 
             1, 1, 0, 0, 1, 0, 1, 0, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 
             1, 1, 1, 0, 0, 0, 0, 0, 1, 1, 0, 1, 1, 0, 0, 0, 1, 0, 1, 0, 1, 0, 0, 0, 
             0, 1, 0, 0, 0, 1, 1, 0, 0, 0, 1, 1, 0, 1, 0, 0, 0, 1, 1, 1, 1, 0, 0, 0, 
             1, 1, 1, 1, 1, 1, 1, 1, 1, 0, 1, 0, 0, 1, 1, 1, 1, 1, 0, 0, 1, 1, 1, 1, 
             0, 0, 0, 1, 1, 1, 0, 1, 1, 1, 0, 1, 0, 0, 1, 1, 0, 1, 0, 0, 1, 1, 0, 0, 
             0, 1, 1, 1, 1, 0, 1, 1, 1, 1, 0, 0, 1, 1, 1, 0, 1, 1, 1, 0, 1, 1),
  
  TimeDays = c(428, 275, 262, 183, 259, 714, 438, 796, 892, 393, 161, 836, 523, 612,
212, 399, 771, 514, 512, 624, 209, 341, 299, 826, 262, 566, 368, 302, 602, 652,
293, 564, 394, 755, 591, 787, 739, 550, 837, 612, 581, 523, 504, 785, 774, 560, 
160, 482, 518, 683, 147, 563, 646, 899, 857, 180, 452, 760, 496, 258, 181, 386, 
439, 563, 337, 613, 192, 405, 667, 905, 247, 821, 821, 517, 346, 294, 244, 95, 376,
212, 96, 532, 522, 679, 408, 840, 148, 168, 489, 541, 205, 475, 237, 517, 749, 150, 
465, 708, 713, 146, 450, 555, 460, 53, 122, 35, 532, 684, 769, 591, 769, 609, 932, 
932, 587, 26, 72, 641, 367, 633, 661, 232, 13, 563, 969, 1052, 944, 881, 190, 79, 884,
170, 286, 358, 326, 769, 161, 564, 268, 611, 322, 1076, 2, 788, 575, 109, 730, 790, 
456, 231, 143, 86, 1021, 684, 878, 216, 808, 268, 222, 683, 496, 389, 126, 17, 350, 
531, 317, 461, 37, 167, 358, 49, 457, 127, 7, 29, 62, 150, 223, 129, 204, 129, 581, 176,
30, 41, 543, 210, 193, 434, 367, 348, 28, 337, 175, 149, 546, 84, 283, 533, 207, 216, 
28, 67, 62, 111, 257, 136, 342, 41, 531, 98, 145, 50, 53, 103, 2, 157, 75, 19, 35, 394, 
117, 175, 180, 314, 480, 325, 280, 204, 366, 531, 59, 33, 540, 551, 90, 47),

  Prison = structure(c(1L, 2L, 1L, 1L, 2L, 1L, 2L, 2L, 1L,
    2L, 2L, 2L, 1L, 1L, 2L, 2L, 2L, 2L, 1L, 2L, 2L, 2L, 1L, 1L,
    2L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 2L, 2L, 1L, 1L, 1L, 2L, 1L,
    1L, 1L, 1L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 1L,
    1L, 2L, 1L, 1L, 1L, 2L, 2L, 1L, 1L, 1L, 1L, 2L, 2L, 1L, 1L,
    1L, 1L, 1L, 2L, 1L, 2L, 1L, 2L, 2L, 2L, 1L, 1L, 1L, 2L, 1L,
    1L, 1L, 2L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 2L, 1L, 2L, 1L,
    1L, 1L, 1L, 1L, 2L, 2L, 2L, 1L, 1L, 2L, 1L, 2L, 2L, 2L, 2L,
    1L, 1L, 2L, 1L, 1L, 1L, 1L, 2L, 2L, 1L, 1L, 1L, 2L, 1L, 2L,
    1L, 2L, 1L, 1L, 1L, 2L, 2L, 1L, 2L, 2L, 2L, 1L, 2L, 2L, 1L,
    1L, 2L, 2L, 1L, 2L, 2L, 2L, 2L, 1L, 2L, 2L, 1L, 1L, 2L, 1L,
    1L, 1L, 1L, 2L, 2L, 1L, 2L, 2L, 2L, 1L, 2L, 1L, 1L, 2L, 1L,
    2L, 2L, 1L, 2L, 2L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 2L, 2L,
    1L, 1L, 2L, 1L, 1L, 2L, 2L, 2L, 1L, 2L, 1L, 2L, 1L, 1L, 2L,
    2L, 1L, 2L, 2L, 1L, 1L, 2L, 1L, 2L, 1L, 1L, 2L, 2L, 2L, 2L,
    2L, 1L, 2L, 1L, 2L, 2L, 1L, 1L, 2L, 1L, 1L, 1L, 2L, 2L, 2L,
    1L, 1L, 1L, 1L),
    .Label = c("No prison record", "Prison record"), class = "factor"),
  
  Dose = c(50, 55, 55, 30, 65, 55, 65,
    60, 50, 65, 80, 60, 55, 70, 60, 60, 75, 80, 80, 80, 60, 60,
    55, 80, 65, 45, 55, 50, 60, 80, 65, 60, 55, 65, 55, 80, 60,
    60, 60, 65, 70, 60, 60, 80, 65, 65, 35, 30, 65, 50, 65, 70,
    60, 60, 60, 70, 60, 60, 65, 40, 60, 60, 80, 75, 65, 60, 80,
    80, 50, 80, 70, 80, 75, 45, 60, 65, 60, 60, 55, 40, 70, 80,
    70, 35, 50, 80, 65, 65, 80, 80, 50, 75, 45, 70, 70, 80, 65,
    60, 50, 50, 55, 80, 50, 60, 60, 40, 70, 65, 70, 70, 40, 100,
    80, 80, 110, 40, 40, 70, 70, 70, 40, 70, 60, 70, 80, 80,
    80, 80, 50, 40, 50, 40, 45, 60, 60, 40, 40, 80, 70, 40, 55,
    80, 40, 70, 80, 70, 80, 90, 70, 60, 70, 40, 80, 80, 60, 100,
    60, 40, 40, 100, 40, 55, 75, 40, 60, 65, 50, 75, 60, 55,
    45, 60, 40, 20, 40, 60, 40, 60, 40, 40, 65, 50, 65, 55, 60,
    60, 40, 50, 70, 55, 45, 60, 50, 40, 60, 80, 50, 45, 80, 55,
    50, 50, 50, 50, 60, 55, 60, 55, 60, 40, 45, 40, 55, 50, 50,
    50, 60, 60, 55, 40, 60, 80, 40, 60, 60, 70, 50, 60, 90, 50,
    55, 50, 45, 60, 80, 65, 40, 45)),

.Names = c("Clinic", "Status", "TimeDays", "Prison", "Dose"),
row.names = c(NA, -238L), class = "data.frame")

head(heroin_dat, n = 10)
```
\pagebreak

## Figure 7.4

```{r, echo=TRUE}
attach(heroin_dat)
plot(survfit(Surv(TimeDays, Status) ~ Clinic), xlab = "Time (days)", ylab = "Proportion",
     lty = c(2,1)) # 1:2
legend("bottomleft", c("Clinic 1", "Clinic 2"), lty = c(2,1)) # 1:2
contrasts(Clinic)
```
\pagebreak

## Table 7.4

```{r, echo=TRUE}
heroin_cox <- coxph(Surv(TimeDays, Status) ~ Prison + Dose + Clinic)
summary(heroin_cox)
```
\pagebreak

Other results mentioned in the text following Table 7.4:

```{r, echo=TRUE}
# Backward elimination etc.
step(heroin_cox, direction="backward")

heroin_zph <- cox.zph(heroin_cox)
heroin_zph

# For Exercise 7.1, use strata as follows:
#heroin_cox <- coxph(Surv(TimeDays, Status) ~ Prison + Dose + strata(Clinic))
```

