---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Examples of Chapter 18:**
  |  **Grouped Multivariate Data**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "20 December 2018"
output: 
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak

# Examples

# Table 18.1: Wechsler Adult Intelligence IQ Scores for Five Men and Five Women

**Source of data**: Willerman, L., Schultz, R., Rutledge, J. N. and Bigler, E. D. (1991). In vivo
brain size and intelligence. *Intelligence*, 15, 223–228. https://doi.org/10.1016/0160-2896(91)90031-8

```{r}
iq_scores <- read.csv("data/iq.csv")
iq_scores[1:5, ]
iq_scores[21:25, ]

n1 = 20; n2 = 20; q = 3
sex <- rep(c(1, 2), c(n1, n2))

# Hotelling's T2 test
m1 <- apply(iq_scores[sex==1, ], 2, mean)
m2 <- apply(iq_scores[sex==2, ], 2, mean)
S1 <- var(iq_scores[sex==1, ])
S2 <- var(iq_scores[sex==2, ])

S <- ((n1-1)*S1 + (n2-1)*S2) / (n1+n2-2)
T2 <- t(m1-m2) %*% solve(S) %*% (m1-m2)
F <- (n1+n2-q-1)*T2/((n1+n2-2)*q)
pvalue <- 1 - pf(F, q, n1+n2-q-1)
c(T2, F, pvalue) # raw output of the above statistics (see below for an alternative!)

```

Quoting our own text on p.367 (here achieved with \LaTeX{} and inline **R** code of **R Markdown**):

Here, Hotelling's $T^2$ takes the value `r round(T2, digits = 2)`, with the corresponding $F$-statistic being `r round(F, 2)`, having 3 and 36 degrees of freedom; the associated $p$-value is `r round(pvalue, 2)`. There is no evidence of a gender difference on the three measures of IQ.

\pagebreak

# Table 18.2: SIDs Data

**Source of data**: Spicer, C. C., Lawrence, C. J. and Southall, D. P. (1987). Statistical analysis
of heart rates and subsequent victims of sudden infant death syndrome.
*Statistics in Medicine*, 6, 159–166. (Appendix II, p.165) https://doi.org/10.1002/sim.4780060208

```{r}
sid <- read.csv("data/sid.csv")
sid[45:54, ]

```
\pagebreak

## Figure 18.2

```{r, out.width='100%', fig.width=6, fig.height=6}
plot(sid$BW, sid$F68, xlab = "Birth Weight (g)", ylab = "Factor 68", type = "n")
text(sid$BW, sid$F68, labels = sid$Group)
legend("topright", c("Controls", "SID victims"), pch = c("1", "2"))
```
\pagebreak

## Table 18.3

```{r}
# Find discriminant function manually using only BW and F68:
sid1 <- sid[sid$Group == 1, -c(1, 2, 5)]
sid2 <- sid[sid$Group == 2, -c(1, 2, 5)]
m1 <- apply(sid1, 2, mean)
m2 <- apply(sid2, 2, mean)
S1 <- var(sid1)
S2 <- var(sid2)
n1 <- length(sid1[, 1])
n2 <- length(sid2[, 1])
S <- ((n1-1)*S1+(n2-1)*S2)/(n1+n2-2)
a <- solve(S)%*%(m1-m2)
z1 <- sum(a*m1)
z2 <- sum(a*m2)
z <- 0.5*(z1+z2)

# results of the Table briefly (could be more verbose):
m1; m2

S1; S2

S

a

z1; z2; z

```
\pagebreak
```{r}
# (mis)classification table:

#install.packages("MASS")
library("MASS")

sid_ldf <- lda(Group ~ BW + F68 + HR + GA, data = sid, prior = c(0.5, 0.5))
table(sid$Group, predict(sid_ldf)$class)

```

So, the (too optimistic) percentage of misclassifications is $(100\times(8+3)/65) \%$ = `r round(100*(8+3)/65, 1)`%.

\pagebreak

## Figure 18.3

```{r, out.width='100%', fig.width=6, fig.height=6}

plot(sid$BW, sid$F68, xlab = "Birth Weight (g)", ylab = "Factor 68", type = "n")
text(sid$BW, sid$F68, labels = sid$Group)
slope <- -a[1]/a[2]
intercept <- z/a[2]
abline(a=intercept, b=slope, lwd=2)
legend("topleft", c("Controls", "SID victims"), pch = c("1", "2"))

```
\pagebreak

# Table 18.4: Data from Investigation of Risk Taking

**Source of data**: Timm, N. H. (2002). *Applied Multivariate Analysis*. Springer, New York. https://doi.org/10.1007/b98963

```{r}
Timm <- read.csv("data/Timm.csv")
str(Timm)
head(Timm, n = 10)
tail(Timm, n = 10)
```
\pagebreak

## Figure 18.4

```{r, fig.width=6, fig.height=6}
par(mfrow=c(2,1))
treatments <- c("AA", "C", "NC")
ylim <- c(0, 80)
with(Timm, {
boxplot(Low[Group==1], Low[Group==2], Low[Group==3], 
        names = treatments, ylab = "'Low' score",  ylim = ylim)
  boxplot(High[Group==1], High[Group==2], High[Group==3], 
        names = treatments, ylab = "'High' score", ylim = ylim)
})

```
\pagebreak

## Table 18.5

```{r}
Timm_manova <- manova(cbind(Low, High) ~ Group, data = Timm)
summary(Timm_manova, test = "Pillai")
summary(Timm_manova, test = "Wilks")
summary(Timm_manova, test = "Hotelling")
summary(Timm_manova, test = "Roy")

```
\pagebreak

## Figure 18.5

```{r, fig.width=5, fig.height=5}
# function copied here from Chapter 12 examples:
chisplot <- function(x) {
    if (!is.matrix(x)) stop("x is not a matrix")
    n <- nrow(x)
    p <- ncol(x)
    xbar <- apply(x, 2, mean)
    S <- var(x)
    S <- solve(S)
    index <- (1 : n) / (n + 1)
    xcent <- t(t(x) - xbar)
    di <- apply(xcent, 1, function(x,S) x %*% S %*% x, S)
    
    quant <- qchisq(index, p)
    plot(quant, sort(di), 
         ylab = "Ordered Distances", 
         xlab = "Chi-square Quantile", 
         lwd = 2, pch = 1)
}

chisplot(residuals(Timm_manova))
abline(a = 0, b = 1)
```
\pagebreak

# Table 18.6: fMRI Data

```{r}
fMRI <- read.csv("data/fMRI.csv")
head(fMRI, n = 20)
tail(fMRI, n = 20)

```
\pagebreak

## Figure 18.6

```{r, fig.width=6, fig.height=6}
plot(fMRI$T2 ~ fMRI$PD, type = "n", pch = fMRI$Class, xlab = "PD", ylab = "T2")
text(fMRI$PD, fMRI$T2, labels = fMRI$Class, cex=0.6)
legend("topleft", c("Grey matter", "White matter", "CSF"), pch = c("1", "2", "3"))

```
\pagebreak

## Table 18.7

```{r, fig.width=6, fig.height=6}
Class <- fMRI$Class
# calculate means of each class
m1 <- apply(fMRI[Class==1, -3], 2, mean)
m2 <- apply(fMRI[Class==2, -3], 2, mean)
m3 <- apply(fMRI[Class==3, -3], 2, mean)

# numbers in each class
n1 <- length(fMRI[Class==1, 1])
n2 <- length(fMRI[Class==2, 1])
n3 <- length(fMRI[Class==3, 1])

# covariance matrices
S1 <- (n1-1)*var(fMRI[Class==1, -3])/(n1-1)
S2 <- (n2-1)*var(fMRI[Class==2, -3])/(n2-1)
S3 <- (n3-1)*var(fMRI[Class==3, -3])/(n3-1)

# pooled covariance matrix
S123 <- ((n1-1)*var(fMRI[Class==1, -3]) + 
         (n2-1)*var(fMRI[Class==2, -3]) +
         (n3-1)*var(fMRI[Class==3,-3])) / (n1+n2+n3-3)

# results briefly:
m1; m2; m3

c(n1, n2, n3)

S1; S2; S3

S123

```
\pagebreak

## Table 18.8

```{r, fig.width=6, fig.height=6}
# coefficients for each classification class
invS <- solve(S123)
a1 <- invS%*%(m1-m2)
a2 <- invS%*%(m1-m3)
a3 <- invS%*%(m2-m3)

# thresholds
z12 <- (m1%*%a1+m2%*%a1)/2
z13 <- (m1%*%a2+m3%*%a2)/2
z23 <- (m2%*%a3+m3%*%a3)/2

# results (very) briefly:

a1; a2; a3

z12; z13; z23

```
\pagebreak

## Figure 18.7

```{r, fig.width=6, fig.height=6}
# code from previous figure:
plot(fMRI$T2 ~ fMRI$PD, type = "n", pch = fMRI$Class, xlab = "PD", ylab = "T2")
text(fMRI$PD, fMRI$T2, labels = fMRI$Class, cex=0.6)
legend("topleft", c("Grey matter", "White matter", "CSF"), pch = c("1", "2", "3"))

# add discriminant functions
abline(z12/a1[2], -a1[1]/a1[2])
abline(z13/a2[2], -a2[1]/a2[2], lty=2)
abline(z23/a3[2], -a3[1]/a3[2], lty=3)
legend("bottomright", c("Grey/White", "Grey/CSF", "White/CSF"), lty=1:3)

```
