---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Examples of Chapter 4:**
  |  **Multiple Linear Regression**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "9 October 2018"
output: 
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak

# Examples

# Figure 4.1: Time Spent Looking After Car

```{r, echo=TRUE}
Gender <- c(1, 1, 0, 1, 0, 0, 0, 1, 1, 0, 0, 1, 1, 1, 0, 1, 0, 0, 1, 0, 0,
         0, 1, 0, 0, 0, 1, 0, 1, 1, 1, 0, 1, 0, 0, 0, 1, 1, 1, 1)

Age <- c(55, 43, 57, 26, 22, 32, 26, 29, 40, 30, 34, 44, 49, 22, 34, 47,
         48, 48, 22, 24, 50, 49, 49, 48, 29, 58, 24, 21, 29, 45, 28, 37,
         44, 22, 38, 24, 34, 26, 26, 25)

Extroversion <- c(40, 45, 52, 62, 31, 28, 2, 83, 55, 32, 47, 45, 60, 13, 7, 85,
           38, 61, 26, 3, 29, 60, 47, 18, 16, 36, 24, 12, 32, 46, 26, 40,
           46, 44, 3, 25, 43, 41, 42, 36)

Time <- c(46, 79, 33, 63, 20, 18, 11, 97, 63, 46, 21, 71, 59, 44, 30, 80,
          45, 26, 33, 7, 50, 54, 73, 19, 36, 31, 71, 15, 40, 61, 45, 42,
          57, 34, 26, 47, 42, 44, 59, 27)

carwash <- as.data.frame(cbind(Gender, Age, Extroversion, Time))
# Time, Age, Extroversion - points by Gender
pairs(carwash[, c(4,2,3)], panel=function(x,y) text(x,y,Gender))
```
\pagebreak

## Table 4.1

```{r, echo=TRUE}
summary(lm(Time ~ Age + Extroversion + Gender, data = carwash))
```
\pagebreak

## Table 4.2

```{r, echo=TRUE}
summary(lm(Time ~ Extroversion + Gender, data = carwash))
```
\pagebreak

## Figure 4.2

```{r, echo=TRUE}
#par(mfrow = c(1,1))
gender <- rep("m", length(Gender))
gender[Gender==0] <- "f"
x <- seq(0, 90, length = 50)
ym <- 15.68 + 19.18 + 0.51*x
yf <- 15.68 + 0.51*x
plot(Time ~ Extroversion, xlab = "Extroversion score",
     ylab = "Time spent looking after car (mins)", type = "n")
text(Extroversion, Time, labels = gender)
lines(x, ym, lty = 1)
lines(x, yf, lty = 2)
legend("topleft", c("Male", "Female"), lty = 1:2)
```
\pagebreak

## Table 4.3

```{r, echo=TRUE}
summary(lm(Time ~ Extroversion + Gender + Extroversion:Gender, data = carwash))
```
\pagebreak

## Figure 4.3

```{r, echo=TRUE}
gender <- rep("m", length(Gender))
gender[Gender==0] <- "f"
x <- seq(0, 90, length = 50)
ym <- 20 + 7.82 + (0.36+0.31)*x
yf <- 20 + 0.36*x
plot(Time ~ Extroversion, xlab = "Extroversion score",
     ylab = "Time spent looking after car (mins)", type = "n")
text(Extroversion, Time, labels = gender)
lines(x, ym, lty = 1)
lines(x, yf, lty = 2)
legend("topleft", c("Male", "Female"), lty = 1:2)
```
\pagebreak

# Table 4.4: Course Evaluation Data

```{r, echo=TRUE}
Overall <- c(3.4, 2.9, 2.6, 3.8, 3.0, 2.5, 3.9, 4.3, 3.8, 3.4,
             2.8, 2.9, 4.1, 2.7, 3.9, 4.1, 4.2, 3.1, 4.1, 3.6,
             4.3, 4.0, 2.1, 3.8, 2.7, 4.4, 3.1, 3.6, 3.9, 2.9,
             3.7, 2.8, 3.3, 3.7, 4.2, 2.9, 3.9, 3.5, 3.8, 4.0,
             3.1, 4.2, 3.0, 4.8, 3.0, 4.4, 4.4, 3.4, 4.0, 3.5)

Teach <- c(3.8, 2.8, 2.2, 3.5, 3.2, 2.7, 4.1, 4.2, 3.7, 3.7,
           3.3, 3.3, 4.1, 3.1, 2.9, 4.5, 4.3, 3.7, 4.2, 4.0,
           3.7, 4.0, 2.9, 4.0, 3.3, 4.4, 3.4, 3.8, 3.7, 3.1,
           3.8, 3.2, 3.5, 3.8, 4.4, 3.7, 4.0, 3.4, 3.2, 3.8,
           3.7, 4.3, 3.4, 4.0, 3.1, 4.5, 4.8, 3.4, 4.2, 3.4)

Exam <- c(3.8, 3.2, 1.9, 3.5, 2.8, 3.8, 3.8, 4.1, 3.6, 3.6,
          3.5, 3.3, 3.6, 3.8, 3.8, 4.2, 4.1, 4.0, 4.3, 4.2,
          4.0, 4.1, 2.7, 4.4, 4.4, 4.3, 3.6, 4.1, 4.2, 3.6,
          4.4, 3.4, 3.2, 3.7, 4.3, 4.1, 3.7, 4.0, 3.6, 4.0,
          3.7, 4.2, 4.2, 4.1, 3.2, 4.5, 4.3, 3.6, 4.0, 3.9)

Knowledge <- c(4.5, 3.8, 3.9, 4.1, 3.5, 4.2, 4.5, 4.7, 4.1, 4.1,
               3.9, 3.9, 4.0, 4.1, 4.5, 4.5, 4.5, 4.5, 4.7, 4.0,
               4.5, 4.6, 3.7, 4.1, 3.6, 4.4, 3.3, 3.8, 4.2, 3.8,
               4.0, 3.1, 4.4, 4.3, 5.0, 4.2, 4.5, 4.5, 4.7, 4.3,
               4.0, 4.2, 3.8, 4.9, 3.7, 4.6, 4.3, 3.5, 4.4, 4.4)

Grade <- c(3.5, 3.2, 2.8, 3.3, 3.2, 3.2, 3.6, 4.0, 3.0, 3.1,
           3.0, 3.3, 3.2, 3.4, 3.7, 3.8, 3.8, 3.7, 4.2, 3.8,
           3.3, 3.2, 3.1, 3.9, 4.3, 2.9, 3.2, 3.5, 3.3, 3.2,
           4.1, 3.5, 3.6, 3.7, 3.3, 3.6, 3.5, 3.4, 3.0, 3.4,
           3.7, 3.8, 3.7, 3.7, 3.3, 4.0, 3.6, 3.3, 4.1, 3.3)

Enroll <- c(21, 50, 800, 221, 7, 108, 54, 99, 52, 47, 73, 25,
            37, 83, 70, 16, 14, 12, 20, 18, 260, 100, 118, 35,
            32, 25, 55, 28, 28, 27, 25, 50, 76, 28, 85, 75, 90,
            94, 65, 100, 105, 70, 49, 64, 700, 27, 15, 40, 18, 90)

courses <- as.data.frame(cbind(Overall, Teach, Exam, Knowledge, Grade, Enroll))
head(courses)
```
\pagebreak

## Figure 4.4

```{r, echo=TRUE}
pairs(courses, panel=function(x,y) {
  points(x,y)
	abline(lm(y ~ x))
	lines(lowess(y ~ x), lty = 3)
	}
)
```
\pagebreak

## Table 4.5

```{r, echo=TRUE}
reg <- lm(Overall ~ Teach + Exam + Knowledge + Grade + Enroll)
summary(reg)
```
\pagebreak
```{r, echo=TRUE}
# remove outliers
reg_345 <- lm(Overall[-c(3, 45)] ~ Teach[-c(3, 45)] + Exam[-c(3, 45)] +
                Knowledge[-c(3, 45)] + Grade[-c(3, 45)] + Enroll[-c(3, 45)])
summary(reg_345)

# VIFs
s1 <- summary(lm(Teach ~ Exam + Knowledge + Grade + Enroll))
s2 <- summary(lm(Exam ~ Teach + Knowledge + Grade + Enroll))
s3 <- summary(lm(Knowledge ~ Exam + Teach + Grade + Enroll))
s4 <- summary(lm(Grade ~ Exam + Knowledge + Teach + Enroll))
s5 <- summary(lm(Enroll ~ Exam + Knowledge + Grade + Teach))
#
rsq <- c(s1$r.squared, s2$r.squared, s3$r.squared, s4$r.squared, s5$r.squared)
rsq
# rounded values used in the book (no difference in practice):
# rsq <- c(0.58, 0.68, 0.33, 0.38, 0.35)
1/(1 - rsq)
```
\pagebreak

## Table 4.6

```{r, echo=TRUE}
step(reg, direction = "backward")
```
\pagebreak

```{r, echo=TRUE}
final_reg <- lm(Overall ~ Teach + Knowledge)
summary(final_reg)
```
\pagebreak

## Figure 4.5

```{r, echo=TRUE}
rawresid <- final_reg$residuals
infl.reg <- lm.influence(final_reg)

r.std <- rawresid/sqrt(0.3202^2 * (1 - infl.reg$hat))
r.del <- rawresid/sqrt(infl.reg$sigma^2 * (1 - infl.reg$hat))

par(mfrow=c(2,2))
boxplot(r.std, ylab = "Standardised residual")
qqnorm(r.std, main = "")
boxplot(r.del, ylab = "Deletion residual")
qqnorm(r.del, main = "")
```
\pagebreak

## Figure 4.6

```{r, echo=TRUE}
par(mfrow = c(3,1), mar = c(4, 4, 2, 0.1))
lab <- "Standardised residual"
plot(predict(final_reg), r.std, xlab = "Fitted Value of Overall", ylab = lab)
plot(Teach, r.std, xlab = "Teacher rating",ylab = lab)
plot(Knowledge, r.std, xlab = "Knowledge of Subject Rating",ylab = lab)
```
\pagebreak

## Figure 4.7

```{r, echo=TRUE}
par(mfrow = c(3,1), mar = c(4, 4, 2, 0.1))
lab <- "Deletion residual"
plot(predict(final_reg), r.del, xlab = "Fitted Value of Overall", ylab = lab)
plot(Teach, r.del, xlab = "Teacher rating",ylab = lab)
plot(Knowledge, r.del, xlab = "Knowledge of Subject Rating",ylab = lab)
```
\pagebreak

## Figure 4.8

```{r, echo=TRUE}
par(mfrow = c(1,1))
plot(1:50, infl.reg$hat, xlab = "Observation Number", ylab = "Cook's Distance")
segments(1:50, rep(0,50), 1:50, infl.reg$hat)
```
\pagebreak

# Table 4.7: Fecundity of Fruit Flies

```{r, echo=TRUE}
flies <- read.table("data/fruitfly.txt", header = T)
flies
```
\pagebreak

## Table 4.8

```{r, echo=TRUE}
n <- 25
x1 <- c(rep(1,n), rep(0,n), rep(-1,n))
x2 <- c(rep(0,n), rep(1,n), rep(-1,n))
response <- c(flies$RS, flies$SS, flies$NS)

flies2 <- cbind(x1, x2, response)
head(flies2)
tail(flies2)

reg0 <- lm(response ~ 1)

reg <- lm(response ~ x1 + x2)
summary(reg)
# Residual sum of squares:
sum(reg$residuals^2)

anova(reg0, reg)
```
\pagebreak

# Table 4.9: A Balanced 2 x 2 Data Set

```{r, echo=TRUE}
x1 <- c(rep(1,8), rep(-1,8))
x2 <- c(rep(1,4), rep(-1,4), rep(1,4), rep(-1,4))
response <- c(23, 25, 27, 29, 26, 32, 30, 31, 22, 23, 21, 21, 37, 38, 40, 35)

cbind(x1, x2, response)

reg0 <- lm(response ~ 1)
reg1 <- lm(response ~ x1)
summary(reg1)
# Residual sum of squares:
sum(reg1$residuals^2)
anova(reg0, reg1)
#
reg2 <- lm(response ~ x1 + x2)
summary(reg2)
sum(reg2$residuals^2)
anova(reg1, reg2)
#
reg3 <- lm(response ~ x1 + x2 + x1:x2)
summary(reg3)
sum(reg3$residuals^2)
anova(reg2, reg3)
```
\pagebreak

# Table 4.10: Unbalanced 2 x 2 Data Set

```{r, echo=TRUE}
x1 <- c(rep(1,21), rep(-1,21))
x2 <- c(rep(1,8), rep(-1,13), rep(1,8), rep(-1,13))
response <- c(23, 25, 27, 29, 30, 27, 23, 25,
              26, 32, 30, 31, NA, NA, NA, NA, NA, NA, NA, NA, NA,
              22, 23, 21, 21, 19, 23, 17, NA, 37, 38, 40, 35, 39,
              35, 38, 41, 32, 36, 40, 41, 38)

cbind(x1, x2, response)

reg0 <- lm(response ~ 1)
reg1 <- lm(response ~ x1)
summary(reg1)
# Residual sum of squares:
sum(reg1$residuals^2)
anova(reg0, reg1)
#
reg2 <- lm(response ~ x1 + x2)
summary(reg2)
sum(reg2$residuals^2)
anova(reg1, reg2)
#
reg3 <- lm(response ~ x1 + x2 + x1:x2)
summary(reg3)
sum(reg3$residuals^2)
anova(reg2, reg3)
```
\pagebreak

