---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Examples of Chapter 10:**
  |  **Analysis of Longitudinal Data III: Non-Normal Responses**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "22 November 2018"
output: 
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak

# Examples

# Beat the Blues Revisited

Using the Beat the Blues data introduced in **Chapter 8**:

```{r, echo=TRUE}
library(tidyr); library(dplyr);

BtB <- read.table("data/BtB.txt", header = TRUE, sep = '\t')

# to make sure that the factor levels are logical (esp. Treatment):
BtB <- within(BtB, {
       Drug <- factor(Drug, levels=c("No", "Yes")) # default
     Length <- factor(Length, levels=c("<6m", ">6m")) # default
  Treatment <- factor(Treatment, levels=c("TAU", "BtheB")) # NOT default!
})

glimpse(BtB); head(BtB); tail(BtB)

# Convert data to long form for the analyses, adding Time:
BtBL <- gather(BtB, key = Visit, value = BDI, BDI2m, BDI4m, BDI6m, BDI8m) %>%
  mutate(Time = as.integer(substr(Visit, 4, 4))) 

glimpse(BtBL); head(BtBL); tail(BtBL)

# important to order the data with respect to the subjects:
BtBLSORT <- BtBL %>% 
  arrange(Subject)

head(BtBLSORT); tail(BtBLSORT)
```
\pagebreak

## Table 10.3

```{r, echo=TRUE}
# install.packages("gee")
library("gee")

BtB_gee1 <- gee(BDI ~ BDIpre + Time + Treatment + Drug + Length, id = Subject, 
                data = BtBLSORT, family = gaussian, corstr = "independence")
summary(BtB_gee1)
```
\pagebreak

## Table 10.4

```{r, echo=TRUE}
BtB_gee2 <- gee(BDI ~ BDIpre + Time + Treatment + Drug + Length, id = Subject,
                data = BtBLSORT, family = gaussian, corstr = "exchangeable")
summary(BtB_gee2)
```
\pagebreak

# Table 10.1: Respiratory Disorder Data

```{r, echo=TRUE}
library(tidyr); library(dplyr); 
# install.packages("HSAUR3")
library("HSAUR3")
data("respiratory", package = "HSAUR3")
RESPL <- respiratory # (data are readily given in long form)
glimpse(RESPL); head(RESPL, n = 5)

RESPL <- within(RESPL, {
     levels(gender) <- c("male", "female") # see the original article by Davis (1991)
     status <- as.numeric(status) - 1 # for converting to wide form below
})
head(RESPL, n = 5)

# convert data to WIDE form:
# (sep: set valid names month0:month4 here; otherwise just 0:4)
RESP <- RESPL %>% 
  spread(key = month, value = status, sep = "") %>% 
  arrange(subject)

glimpse(RESP); head(RESP, n = 5)

# back to long form; leave month0 for baseline:
RESPL <- gather(RESP, key = months, value = status, month1:month4) %>%
  mutate(month = as.integer(substr(months, 6, 6))) %>% 
  rename(baseline = month0) %>% 
  arrange(subject)

glimpse(RESPL); head(RESPL, n = 5)
```
\pagebreak

## Figure 10.2

```{r, echo=TRUE}
n = dim(RESP)[1]
RP <- RESP %>%
  group_by(treatment) %>%
  summarise(m0 = 100*sum(month0)/n, 
            m1 = 100*sum(month1)/n, 
            m2 = 100*sum(month2)/n, 
            m3 = 100*sum(month3)/n, 
            m4 = 100*sum(month4)/n ) %>%
  select(-treatment) %>%
  ungroup()

RP2 <- cbind(RP[1, ], RP[2, ])
month <- rep(1:5, times = 2)
treatment <- rep(0:1, each = 5)
RPos <- as.data.frame(cbind(treatment, month, t(RP2)))
row.names(RPos) <- 1:10
names(RPos) <- c("treatment", "month", "positive")
RPos$treatment <- factor(RPos$treatment, labels = c("placebo", "treatment"))

library(ggplot2)

p1 <- ggplot(RPos, aes(x = month, y = positive, fill = treatment))
p2 <- p1 + geom_bar(position = "dodge", stat = "identity")
p3 <- p2 + scale_x_continuous(name = "", breaks = 1:5, 
                labels = c("Baseline", "Month 1", "Month 2", "Month 3", "Month 4"))
p4 <- p3 + theme_bw() + theme(panel.grid.major = element_blank(), 
                              panel.grid.minor = element_blank())
p5 <- p4 + scale_y_continuous(name = "Positive responses (%)")
p6 <- p5 + scale_fill_grey(start = 0.5, end = 0)
p6
```
\pagebreak

## Table 10.5

```{r, echo=TRUE}
resp_gee1 <- gee(status ~ age, data = RESPL, family = binomial, id = subject, 
                 corstr = "independence", scale.fix = TRUE, scale.value = 1)
summary(resp_gee1)
```
\pagebreak

## Table 10.6

```{r, echo=TRUE}
resp_gee2 <- gee(status ~ age, data = RESPL, family = binomial, id = subject,
                 corstr = "exchangeable", scale.fix = TRUE, scale.value = 1)
summary(resp_gee2)
```
\pagebreak

## Table 10.7

```{r, echo=TRUE}
formula3 <- status ~ age + month + baseline + centre + treatment + gender
resp_gee3 <- gee(formula3, data = RESPL, family = binomial, id = subject, 
                 corstr = "exchangeable", scale.fix = TRUE, scale.value = 1)
summary(resp_gee3)
```
\pagebreak

## Table 10.10

```{r, echo=TRUE}
# install.packages("lme4")
library("lme4")

formula4 <- status ~ age + month + baseline + centre + treatment + gender + (1 | subject)
resp_glmer <- glmer(formula4, data = RESPL, family = binomial)
summary(resp_glmer)
```
\pagebreak

# Table 10.2: Data from a Clinical Trial of Patients Suffering from Epilepsy

```{r, echo=TRUE}
library("HSAUR3")
EPIL <- epilepsy # (data are readily given in long form)
glimpse(EPIL); head(EPIL)

# convert data to WIDE form:
# (sep: set valid names month0:month4 here! otherwise just 0:4)

EPI <- EPIL %>% 
  spread(key = period, value = seizure.rate, sep = "") %>% 
  arrange(subject)

glimpse(EPI); head(EPI)

# correct the baseline when converting to long form:

EPIL <- gather(EPI, key = periods, value = seizure.rate, period1:period4) %>%
  mutate( base2 = base/4, week = as.integer(substr(periods, 7, 7)) ) %>% 
  arrange(subject)

glimpse(EPIL); head(EPIL)
```
\pagebreak

## Figure 10.3

```{r, echo=TRUE}
# Convert data to long form, including the baseline;
# correct the baseline first:
EPI0 <- EPI %>% 
  mutate(base2 = base/4)

EPIL0 <- gather(EPI0, key = periods, value = seizure.rate, base2, period1:period4) %>% 
  arrange(subject)

glimpse(EPIL0); head(EPIL0)

# Mark some of the outliers:
# https://stackoverflow.com/questions/33524669/labeling-outliers-of-boxplots-in-r
is_outlier <- function(x) {
  # changed the upper coeff from 1.5 to 3.5 (only more extreme observations highlighted)
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 3.5 * IQR(x))
}
# use subject id (not seizure.rate itself!) for identification:
EPIL0 <- EPIL0 %>%
  mutate(outlier = ifelse(is_outlier(seizure.rate), subject, as.numeric(NA)))

p1 <- ggplot(EPIL0, aes(x = factor(periods), y = seizure.rate, fill = treatment))
p2 <- p1 + geom_boxplot() + geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.6)
p3 <- p2 + facet_grid(treatment ~ ., labeller = label_parsed)
p4 <- p3 + theme_bw() + theme(panel.grid.major = element_blank(), 
                              panel.grid.minor = element_blank())
p5 <- p4 + theme(legend.position = "none")
p6 <- p5 + scale_x_discrete(name = "", 
                            labels = c("Baseline", "Time 1", "Time 2", "Time 3", "Time 4"))
p7 <- p6 + scale_y_continuous(name = "Seizure rate (count)")
p8 <- p7 + scale_fill_grey(start = 1, end = 1) # B&W for the book
p7; p8
```

## Table 10.8

```{r, echo=TRUE}
t1 <- tapply(EPIL0$seizure.rate, EPIL0$period, mean)
t2 <- tapply(EPIL0$seizure.rate, EPIL0$period, var)
t1 <- as.data.frame(t1)
t2 <- as.data.frame(t2)

tab108 <- cbind(t1, t2)
names(tab108) = c("mean", "variance")
tab108
```
\pagebreak

## Table 10.9

```{r, echo=TRUE}
formula5 <- seizure.rate ~ week + age + log(base2) + treatment
epil_gee1 <- gee(formula5, data = EPIL, family = "poisson", id = subject, 
                 corstr = "exchangeable", scale.fix = FALSE)
summary(epil_gee1)
```
\pagebreak

## Table 10.11

```{r, echo=TRUE}
formula6 <- seizure.rate ~ week + age + log(base2) + treatment + (1 | subject)
epil_glmm <- glmer(formula6, data = EPIL, family = "poisson")
summary(epil_glmm)
```
\pagebreak

## Table 10.12

```{r, echo=TRUE}
formula7 <- seizure.rate ~ week + age + log(base2) + treatment + (1 + week | subject)
epil_glmm2 <- glmer(formula7, data = EPIL, family = "poisson")
summary(epil_glmm2)
```
