---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Examples of Chapter 16:**
  |  **Confirmatory Factor Analysis and Structural Equation Models**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "14 December 2018"
output: 
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak

# Examples

# Table 16.1: Observed Correlations for the Ability and Aspiration Example

```{r}
CORR <- matrix(c(1.00,0.00,0.00,0.00,0.00,0.00,
                 0.73,1.00,0.00,0.00,0.00,0.00,
                 0.70,0.68,1.00,0.00,0.00,0.00,
                 0.58,0.61,0.57,1.00,0.00,0.00,
                 0.46,0.43,0.40,0.37,1.00,0.00,
                 0.56,0.52,0.48,0.41,0.72,1.00), 
               ncol = 6, byrow = TRUE)
varNames <- c("SCA", "PPE", "PTE", "PFE", "EA", "CP")
dimnames(CORR) <- list(varNames, varNames)
CORR
```
\pagebreak

## Figure 16.1

```{r}
#install.packages("sem")
#install.packages("semPlot")
library(sem)
library(semPlot)

AAmodel <-specifyModel(text = "
   Ability     -> SCA,        lambda[1], NA
   Ability     -> PPE,        lambda[2], NA
   Ability     -> PTE,        lambda[3], NA
   Ability     -> PFE,        lambda[4], NA
   Aspiration  -> EA,         lambda[5], NA
   Aspiration  -> CP,         lambda[6], NA
   Ability    <-> Aspiration, rho,       NA
   SCA        <-> SCA,        u[1],      NA
   PPE        <-> PPE,        u[2],      NA
   PTE        <-> PTE,        u[3],      NA
   PFE        <-> PFE,        u[4],      NA
   EA         <-> EA,         u[5],      NA
   CP         <-> CP,         u[6],      NA
   Ability    <-> Ability,    NA,        1
   Aspiration <-> Aspiration, NA,        1
  ")

options(fit.indices = c("CFI", "NNFI", "RMSEA", "SRMR")) # (NNFI = TLI)

AAsem <- sem(AAmodel, CORR, N = 556)

semPaths(AAsem, # filetype = "pdf", filename = "AAmodel-KV",
         what = "path", whatLabels = "name", style = "lisrel", layout = "tree2", 
         intercepts = FALSE, residuals = TRUE, thresholds = FALSE, reorder = FALSE,
         rotation = 1, nCharNodes = 0, nCharEdges = 0, 
         sizeMan = 8, sizeLat = 20,  sizeMan2 = 6, sizeLat2 = 10,
         manifests = varNames, latents = c("Ability", "Aspiration"), 
         residScale = 12, as.expression = c("edges", "nodes"), centerLevels = FALSE, 
         edge.label.cex = 1.2, esize = 2.5, label.scale = TRUE, curvePivot = TRUE, 
         curvePivotShape = 0.67, edge.label.position = 0.67, width = 6, height = 2.5)
```
\pagebreak

## Table 16.2

```{r}
summary(AAsem)
```
\pagebreak

# Figure 16.2: Path diagram for the drug usage model

Revisiting the drug usage example introduced in **Chapter 13**:

```{r, out.width='100%'}
DRUGcorr <- matrix(c(
  1.0,rep(0,12),
  0.447,1.0,rep(0,11),
  0.422,0.619,1.0,rep(0,10),
  0.436,0.604,0.583,1.0,rep(0,9),
  0.114,0.068,0.053,0.115,1.0,rep(0,8),
  0.203,0.146,0.139,0.256,0.349,1.0,rep(0,7),
  0.091,0.103,0.110,0.122,0.209,0.221,1.0,rep(0,6),
  0.082,0.063,0.066,0.097,0.321,0.355,0.201,1.0,rep(0,5),
  0.513,0.445,0.365,0.482,0.186,0.316,0.150,0.154,1.0,rep(0,4),
  0.304,0.318,0.240,0.368,0.303,0.377,0.163,0.219,0.534,1.0,rep(0,3),
  0.245,0.203,0.183,0.255,0.272,0.323,0.310,0.288,0.301,0.302,1.0,rep(0,2),
  0.101,0.088,0.074,0.139,0.279,0.367,0.232,0.320,0.204,0.368,0.340,1.0,0.0,
  0.245,0.199,0.184,0.293,0.278,0.545,0.232,0.314,0.394,0.467,0.392,0.511,1.0), ncol=13, byrow=T)
DRUGnames <- c("Cigs", "Beer", "Wine", "Liqr", "Cocn", "Tran", "Drug", "Hern", 
               "Marj", "Hash", "Inhl", "Hall", "Amph")
dimnames(DRUGcorr) <- list(DRUGnames, DRUGnames)
DRUGcorr <- DRUGcorr + t(DRUGcorr)
diag(DRUGcorr) <- diag(DRUGcorr) - 1
DRUGcorr

DRUGmodel <-specifyModel(text = "

Alcohol   -> Cigs,  lambda[1],NA
Cannabis  -> Cigs,  lambda[2],NA
Alcohol   -> Beer,  lambda[3],NA
Alcohol   -> Wine,  lambda[4],NA
Cannabis  -> Wine,  lambda[5],NA
Alcohol   -> Liqr,  lambda[6],NA
HardDrug  -> Liqr,  lambda[7],NA
HardDrug  -> Cocn,  lambda[8],NA
HardDrug  -> Tran,  lambda[9],NA
HardDrug  -> Drug,  lambda[10],NA
HardDrug  -> Hern,  lambda[11],NA
Cannabis  -> Marj,  lambda[12],NA
Cannabis  -> Hash,  lambda[13],NA
HardDrug  -> Hash,  lambda[14],NA
HardDrug  -> Inhl,  lambda[15],NA
HardDrug  -> Hall,  lambda[16],NA
HardDrug  -> Amph,  lambda[17],NA

Cigs     <-> Cigs,  u[1],NA
Beer     <-> Beer,  u[2],NA
Wine     <-> Wine,  u[3],NA
Liqr     <-> Liqr,  u[4],NA
Cocn     <-> Cocn,  u[5],NA
Tran     <-> Tran,  u[6],NA
Drug     <-> Drug,  u[7],NA
Hern     <-> Hern,  u[8],NA
Marj     <-> Marj,  u[9],NA
Hash     <-> Hash,  u[10],NA
Inhl     <-> Inhl,  u[11],NA
Hall     <-> Hall,  u[12],NA
Amph     <-> Amph,  u[13],NA

Alcohol  <-> Cannabis, rho[1],NA
Alcohol  <-> HardDrug, rho[2],NA
Cannabis <-> HardDrug, rho[3],NA

Alcohol  <-> Alcohol, NA,1
Cannabis <-> Cannabis,NA,1
HardDrug <-> HardDrug,NA,1

")

options(fit.indices = c("CFI", "NNFI", "RMSEA", "SRMR")) # (NNFI = TLI)

DRUGsem <- sem(DRUGmodel, DRUGcorr, N = 1634)
```
\pagebreak
```{r, out.width='100%', fig.width=6, fig.height=6}
semPaths(DRUGsem, # filetype = "pdf", filename = "DRUGmodel-KV",
         what = "path", whatLabels = "hide", 
         style = "lisrel", layout = "tree2", 
         residuals = TRUE, rotation = 2, nCharNodes = 0, 
         sizeMan = 8, sizeLat = 10, sizeMan2 = 4, sizeLat2 = 10, 
         esize = 2, label.cex = 0.75, label.scale = FALSE, width = 6, height = 5)

```
\pagebreak

## Table 16.3

```{r}
summary(DRUGsem)
```
\pagebreak

# Figure 16.3: Corrgram for the Systems Intelligence data

**Source of data**: Törmänen, J., Hämäläinen, R. P. and Saarinen, E. (2016). Systems Intelligence
inventory. *The Learning Organization*, 23, 218–231. (The data and parts of the R code by *Juha Törmänen* are used below with kind permission from the authors.)

**See also**:

- Saarinen, E. and Hämäläinen, R. P. (2004). Systems intelligence: Connecting engineering thinking with human sensitivity. In *Systems Intelligence—Discovering a Hidden Competence in Human Action and Organizational Life* (eds. R. P. Hämäläinen and E. Saarinen). Helsinki University of Technology, Research Reports A88. http://sal.aalto.fi/publications/pdffiles/systemsintelligence2004.pdf.

- Hämäläinen, R. P., Jones, R. and Saarinen, E. (2014). *Being Better Better: Living with Systems Intelligence.* Aalto University Publications Crossover 4/2014, http://systemsintelligence.aalto.fi/being_better_better/. 

```{r}
# Set seed to remove effect of random generator
set.seed(42)

# Load data (sidata.csv contains renamed answer columns with names si1-si32)
si_cfa <- list()
si_cfa$data <- read.csv2('data/sidata.csv')
si_cfa$dataset <- si_cfa$data$set

# SI model answers for the 32 items
si_cfa$answers <- si_cfa$data[, 10:41]

head(si_cfa$answers)
```
\pagebreak
```{r, out.width='100%', fig.width=6, fig.height=6}
#install.packages("corrplot")
library(corrplot)

mcor <- cor(si_cfa$answers, use = "complete.obs")
# Add leading zeros to var names for the "alphabet" option of corrplot:
rownames(mcor) <- c("si01", "si02", "si03", "si17", "si13", "si14", "si15", "si04",
                    "si25", "si05", "si26", "si06", "si27", "si28", "si07", "si18", 
                    "si20", "si21", "si09", "si10", "si11", "si22", "si08", "si23", 
                    "si19", "si16", "si29", "si30", "si12", "si24", "si31", "si32")
colnames(mcor) <- rownames(mcor)
corrplot(mcor, method = "shade", order = "alphabet", 
               tl.cex = 0.7, tl.col = "black", tl.srt = 45)

```
\pagebreak

## Figure 16.4

```{r}
# DATA: Validation dataset rows with no missing data (N=815):
si_cfa$data.validation <- si_cfa$data[rowSums(is.na(si_cfa$answers)) == 0 &
                                                    si_cfa$dataset == 'validation', ]

# MODEL: Eight factors, free loadings:
# Free weight model: Each item is explained by one factor, factors have unit variance
# In total 8*4 loadings + 7*8/2 factor covariances + 8*4 item variances = 92 parameters

model.8free <- specifyEquations(text = "

si1  =  a1(1)*PER
si2  =  a2(1)*PER
si3  =  a3(1)*PER
si4  =  a4(1)*PER

si5  =  a5(1)*ATT
si6  =  a6(1)*ATT
si7  =  a7(1)*ATT
si8  =  a8(1)*ATT

# ATD factor starting values coded negatively here; othersize sem has trouble converging
si9  =  a9(1)*ATD
si10 = a10(-1)*ATD
si11 = a11(1)*ATD
si12 = a12(1)*ATD

si13 = a13(1)*DIS
si14 = a14(1)*DIS
si15 = a15(1)*DIS
si16 = a16(1)*DIS

si17 = a17(1)*REF
si18 = a18(1)*REF
si19 = a19(1)*REF
si20 = a20(1)*REF

si21 = a21(1)*WIS
si22 = a22(1)*WIS
si23 = a23(1)*WIS
si24 = a24(1)*WIS

si25 = a25(1)*ENG
si26 = a26(1)*ENG
si27 = a27(1)*ENG
si28 = a28(1)*ENG

si29 = a29(1)*EFF
si30 = a30(-1)*EFF
si31 = a31(1)*EFF
si32 = a32(1)*EFF

V(PER) = 1
V(ATT) = 1
V(ATD) = 1
V(DIS) = 1
V(REF) = 1
V(WIS) = 1
V(ENG) = 1
V(EFF) = 1

", covs = "PER,ATT,ATD,DIS,REF,WIS,ENG,EFF")

options(fit.indices = c("SRMR", "CFI", "RMSEA"))

res.8free <- sem(model = model.8free, data = si_cfa$data.validation, objective = objectiveGLS)
# summary(res.8free) outputs CFI = 0.951, RMSEA = 0.048, SRMR = 0.068

semPaths(res.8free, # filetype = "pdf", filename = "SImodel-KV",
         what = "path", whatLabels = "hide", style = "lisrel",
         residuals = TRUE, curve = 3.2, rotation = 4, nCharNodes = 0,
         sizeLat = 7, sizeMan = 3, sizeMan2 = 2,  esize = 1.3, mar = c(3, 8, 3, 15))
```
\pagebreak
```{r}
summary(res.8free)
```
\pagebreak

# Figure 16.5: Path diagram for the hypothesized structural equation model for job satisfaction data

**Source of data**: Kline, R. B. (2016). *Principles and Practice of Structural Equation Modeling*,
4th edition. Guilford Press, New York.

```{r}
# Modified from https://www.guilford.com/add/kline/houghton-lavaan.r

# Fully latent model of thought strategies and job satisfaction

#install.packages("lavaan")
library(lavaan)

# Input the correlations in lower diagonal form
houghtonLower.cor <- '
1.000
 .668 1.000
 .635  .599 1.000
 .263  .261  .164 1.000
 .290  .315  .247  .486 1.000
 .207  .245  .231  .251  .449 1.000
-.206 -.182 -.195 -.309 -.266 -.142 1.000 
-.280 -.241 -.238 -.344 -.305 -.230  .753 1.000
-.258 -.244 -.185 -.255 -.255 -.215  .554  .587 1.000 
 .080  .096  .094 -.017  .151  .141 -.074 -.111  .016 1.000
 .061  .028 -.035 -.058 -.051 -.003 -.040 -.040 -.018  .284 1.000
 .113  .174  .059  .063  .138  .044 -.119 -.073 -.084  .563  .379 1.000 '

# Name the variables and convert to full correlation matrix
# KV: change names of manifest and factor variables (more according to the article diagram):

houghtonFull.cor <- getCov(houghtonLower.cor, 
         names = c("JW1","JW2","JW3", "UF1","UF2","FOR", "DA1","DA2","DA3", "EBA","ST","MI"))

# add the standard deviations and convert to covariances
houghtonFull.cov <- cor2cov(houghtonFull.cor, 
         sds = c(0.939, 1.017, 0.937, 0.562, 0.760, 0.524, 0.585, 0.609, 0.731, 0.711, 1.124, 1.001))
houghtonFull.cov

# Specify SEM (with Lavaan notation):

JSmodel <- '

# measurement part
CTS =~ EBA + ST + MI
DTP =~ DA1 + DA2 + DA3
SWB =~ UF1 + UF2 + FOR
JS =~ JW1 + JW2 + JW3

# error covariance
UF1 ~~ FOR

# structural part
DTP ~ CTS
SWB ~ CTS + DTP
JS  ~ CTS + DTP + SWB 
'

options(fit.indices = c("CFI", "NNFI", "RMSEA", "SRMR")) # (NNFI = TLI)

# fit JSmodel to data
JSsem <- sem(JSmodel, sample.cov = houghtonFull.cov, sample.nobs = 263)

longLabels = c("JW1","JW2","JW3", "UF1","UF2","FOR", "DA1","DA2","DA3", "EBA","ST","MI",
               "Constructive \n Thought \n Strategies",
               "Dysfunctional \n Thought \n Processes",
               "Subjective \n Well-Being",
               "Job \n Satisfaction")

semPaths(JSsem, # filetype = "pdf", filename = "JSmodel-KV", 
         what = "path", whatLabels = "hide", style = "lisrel", layout = "tree2", 
         residuals = TRUE, nodeLabels = longLabels, nCharNodes = 0, 
         sizeMan = 4.5, sizeLat = 14, sizeLat2 = 10, 
         esize = 2, label.cex = 0.7, label.scale = FALSE, width = 6, height = 5)

summary(JSsem, fit.measures = TRUE, standardized = TRUE, rsquare = TRUE)
```