---
title:  |  
  |  *Multivariate Analysis for the Behavioral Sciences,*
  |  Second Edition (Chapman and Hall/CRC, 2019)
  |  **Examples of Chapter 15:**
  |  **Exploratory Factor Analysis**
author: "Kimmo Vehkalahti and Brian S. Everitt"
date: "12 December 2018"
output: 
  pdf_document:
    number_sections: no
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak

# Examples

# Table 15.1: Maximum Likelihood Two-Factor Solution for Correlations of Six School Subjects

```{r}
R <- c(  1.00, 0.44, 0.41, 0.29, 0.33, 0.25,
         0.44, 1.00, 0.35, 0.35, 0.32, 0.33,
         0.41, 0.35, 1.00, 0.16, 0.19, 0.18,
         0.29, 0.35, 0.16, 1.00, 0.59, 0.47,
         0.33, 0.32, 0.19, 0.59, 1.00, 0.46,
         0.25, 0.33, 0.18, 0.47, 0.46, 1.00)

school_subjects <- c("French", "English", "History", "Arithmetic", "Algebra", "Geometry")
R <- matrix(R, ncol = 6, byrow = TRUE, dimnames = list(school_subjects, school_subjects))
R

Rfa1 <- factanal(covmat = R, factors = 1, n.obs = 220, rotation = "none")
Rfa2 <- factanal(covmat = R, factors = 2, n.obs = 220, rotation = "none")
Rfa3 <- factanal(covmat = R, factors = 3, n.obs = 220, rotation = "none")
Rfa1
Rfa2
Rfa3
```
\pagebreak

```{r}
#install.packages("psych")
library(psych)

RfaNone <- fa(R, nfactors = 2, fm = "ml", rotate = "none")
print.psych(RfaNone)
```

Correlation matrix of the six school subjects implied by the fitted two-factor model:

```{r}
lambdaHat <- as.matrix(RfaNone$loadings)
Rhat <- lambdaHat %*% t(lambdaHat) + RfaNone$residual
Rhat
```
\pagebreak

## Figure 15.1

```{r, fig.width=6, fig.height=6}
plot(RfaNone$loadings, type = "n", 
     xlim = c(-1, +1), ylim = c(-1, +1), 
     xlab = "Factor 1", ylab = "Factor 2")
text(RfaNone$loadings, labels = school_subjects)
abline(h = 0, v = 0)
```
\pagebreak
```{r, fig.width=5, fig.height=5}
# same figure with ggplot (overkill, perhaps) - just testing:
library(ggplot2)
RfaNoneLoadings <- as.matrix(RfaNone$loadings)
class(RfaNoneLoadings) <- "matrix"
RfaNoneLoadings <- as.data.frame(RfaNoneLoadings)
RfaNoneLoadings$Subject <- row.names(RfaNoneLoadings)

p1 <- ggplot(RfaNoneLoadings, aes(x = ML1, y = ML2)) + geom_point()
p2 <- p1 + geom_text(aes(x = ML1+0.03, y = ML2+0.00, 
                         label = RfaNoneLoadings$Subject), size=4, hjust=0)
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), 
                              panel.grid.minor = element_blank())
p4 <- p3 + scale_x_continuous(limits = c(-1, +1), name = "Factor 1")
p5 <- p4 + scale_y_continuous(limits = c(-1, +1), name = "Factor 2")
p6 <- p5 + geom_vline(xintercept=0) + geom_hline(yintercept=0)
p6
```
\pagebreak

## Figure 15.2

```{r, fig.width=7, fig.height=4}
par(mfrow=c(1,2))

plot(RfaNone$loadings, type = "n", xlim = c(0.03, 0.8), ylim = c(-0.5, +0.5), 
     xlab = "Factor 1\n(a)", ylab = "Factor 2")
text(RfaNone$loadings, labels = school_subjects, cex=0.5)
abline(h = 0)
abline(0, +1.5, lwd = 3)
abline(0, -1.5, lwd = 3)

RfaVarimax <- fa(R, nfactors = 2, fm = "ml", rotate = "varimax") # See the note below!

plot(RfaVarimax$loadings, type = "n", xlim = c(0.03, 0.8), ylim = c(0.03, 0.8),
     xlab = "Factor 1\n(b)", ylab = "Factor 2")
text(RfaVarimax$loadings, labels = school_subjects, cex=0.5)
abline(h = 0, v = 0, lwd = 3)
```

Above, the rotation was achieved using the varimax, because, as far as we know, no function for graphical rotation exists in R. For an inspiration to program such a function, you should take a good look at **Survo**: *an environment for creative processing of text and numerical data*, created by professor **Seppo Mustonen**  
(see www.survo.fi/presentation/history.html), cf. the example of *classical music composers* in **Chapter 14**.

**Interactive graphical rotation** has been available in various generations of Survo **since the 1970s**:

* A short demo on YouTube shows how an **oblique graphical rotation** is applied in **SURVO MM**:  
https://youtu.be/-yZRUR6gpAg (see www.survo.fi/demos/#ex38 for more details and links).

* A historical example of using the graphical rotation in **SURVO 76** appears on pp. 29-30 of the report:  
Mustonen, S. (1977). SURVO 76, a statistical data processing system. Research Report No. 6. Dept of Statistics, University of Helsinki. www.survo.fi/publications/Research_Report_6_Mustonen_1977.pdf

Publications related to Survo since 1963: www.survo.fi/publications

\pagebreak

## Table 15.2

```{r}
RfaVarimax <- fa(R, nfactors = 2, fm = "ml", rotate = "varimax")
print.psych(RfaVarimax, sort = TRUE)
```
\pagebreak

## Table 15.3

```{r}
#install.packages("GPArotation")
library(GPArotation)
RfaOblimin <- fa(R, nfactors = 2, fm = "ml", rotate = "oblimin")
print.psych(RfaOblimin, sort = TRUE)
```
\pagebreak

# Table 15.4: Estimated Parameters for the Three-Factor Model Fitted to the Crime Rate Data by Maximum Likelihood

```{r}
# see Chapter 13, where this data set was used for the first time:
crime <- read.table("data/crime.txt", sep = '\t')

crime_fa1 <- factanal(crime, factors = 1, rotation = "none")
crime_fa2 <- factanal(crime, factors = 2, rotation = "none")
crime_fa3 <- factanal(crime, factors = 3, rotation = "none")

crime_fa1
crime_fa2
crime_fa3
```
\pagebreak

```{r}
crime_fa3None <- fa(crime, nfactors = 3, fm = "ml", rotate = "none")
print.psych(crime_fa3None, digits = 3)
```
\pagebreak

## Table 15.5

```{r}
crime_fa3varimax <- fa(crime, nfactors = 3, fm = "ml", rotate = "varimax")
print.psych(crime_fa3varimax, digits = 3, sort = TRUE)
```
\pagebreak

## Figure 15.3

```{r, fig.width=8, fig.height=8}
crime_fa3 <- factanal(crime, factors = 3, method = "mle", rotation = "varimax", 
                      scores = "regression")
pairs(crime_fa3$scores, 
      panel = function(x,y) text(x, y, labels = row.names(crime), cex=0.8))
```